---
title: "Bayesian Prior Justification Report"
subtitle: "`r params$trial_name`"
author: "`r params$author`"
date: "`r params$date`"

params:
  rds_path:    ""
  trial_name:  "Clinical Trial"
  sponsor:     "Sponsor"
  author:      "Biostatistics"
  date:        ""
  notes:       ""

format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    number-sections: true
    theme: flatly
    highlight-style: tango
    code-fold: true
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
  pdf:
    toc: true
    number-sections: true
    pdf-engine: xelatex
    documentclass: article
    latex-auto-install: false
    fig-width: 7
    fig-height: 4
  docx:
    toc: true
    toc-depth: 3
    fig-width: 6.5
    fig-height: 4
    fig-cap-location: bottom

execute:
  echo: false
  warning: false
  message: false

knitr:
  opts_chunk:
    dpi: 150
---

```{=html}
<style>
/* Table striping -- replaces Bootstrap table-striped class lost in knitr path */
table {
  width: 100%;
  border-collapse: collapse;
  margin-bottom: 1rem;
  font-size: 0.9em;
}
table thead tr th {
  background-color: #e9ecef;
  border-bottom: 2px solid #dee2e6;
  padding: 0.5rem 0.75rem;
  text-align: left;
}
table tbody tr:nth-child(even) td {
  background-color: #f8f9fa;
}
table tbody tr td {
  padding: 0.45rem 0.75rem;
  border-top: 1px solid #dee2e6;
}
table caption {
  caption-side: top;
  font-style: italic;
  color: #6c757d;
  margin-bottom: 0.4rem;
  font-size: 0.85em;
}
/* Constrain figure width */
img {
  max-width: 100%;
  height: auto;
}
</style>
```

```{r setup}
if (!requireNamespace("bayprior", quietly = TRUE)) {
  stop(paste0(
    "The 'bayprior' package is not installed in the R library used by Quarto.\n",
    "Run devtools::install() in your R console, then try downloading the report again."
  ))
}
library(bayprior)
library(ggplot2)
library(knitr)

# -- Load R objects from RDS ---------------------------------------------------
# All complex R objects (bayprior, ggplot2) are passed via an .rds file rather
# than execute_params, which serializes to YAML and cannot handle S3 objects.
if (nzchar(params$rds_path) && file.exists(params$rds_path)) {
  .session <- readRDS(params$rds_path)
} else {
  .session <- list(prior=NULL, conflict=NULL, sensitivity=NULL,
                   prior_plot=NULL, overlay_plot=NULL,
                   tornado_plot=NULL, heatmap_plot=NULL,
                   robust_prior=NULL, robust_plot=NULL,
                   sceptical_prior=NULL, sceptical_plot=NULL,
                   power_prior=NULL, power_plot=NULL)
}

pr  <- .session$prior
cd  <- .session$conflict
sa  <- .session$sensitivity
has_prior <- !is.null(pr)
has_conf  <- !is.null(cd)
has_sens  <- !is.null(sa)

pp_prior_plot   <- .session$prior_plot
pp_overlay_plot <- .session$overlay_plot
pp_tornado_plot <- .session$tornado_plot
pp_heatmap_plot <- .session$heatmap_plot

rob_prior     <- .session$robust_prior
pp_rob_plot   <- .session$robust_plot
scep_prior    <- .session$sceptical_prior
pp_scep_plot  <- .session$sceptical_plot
pow_result    <- .session$power_prior
pp_pow_plot   <- .session$power_plot
has_robust    <- !is.null(rob_prior) || !is.null(scep_prior) || !is.null(pow_result)

# -- Format detection ----------------------------------------------------------
# Callout blocks require tcolorbox in PDF but work in HTML and Word.
# Use plain blockquotes for PDF, rich callouts for other formats.
.fmt <- knitr::pandoc_to()
.is_pdf  <- isTRUE(grepl("latex|pdf|beamer", .fmt))
.is_docx <- isTRUE(grepl("docx", .fmt))

.callout <- function(type, text) {
  if (.is_pdf) {
    prefix <- switch(type,
      note      = "**Note:**",
      warning   = "**Warning:**",
      important = "**Important:**",
      tip       = "**Tip:**",
      ""
    )
    # Markdown blockquotes require every line prefixed with ">", and a
    # list must start on its own line (not share a line with preceding
    # text) or pandoc collapses it into a run-on paragraph instead of a
    # real list -- verified empirically against pandoc's actual output.
    body  <- if (nzchar(prefix)) paste0(prefix, "\n\n", text) else text
    lines <- strsplit(body, "\n", fixed = TRUE)[[1]]
    lines[lines == ""] <- " "  # blank quoted lines need a space after ">"
    cat(paste0(paste0("> ", lines, collapse = "\n"), "\n\n"))
  } else {
    cat(paste0("::: {.callout-", type, "}\n", text, "\n:::\n\n"))
  }
}

# -- Status badges (colour-coded cells) -----------------------------------
# kableExtra::cell_spec() supports HTML and LaTeX/PDF, but NOT Word/docx --
# pandoc does not process its embedded HTML/LaTeX styling when converting
# to docx. Rather than silently rendering nothing (or raw markup) in Word,
# fall back to a plain-text bracketed tag there so the flag is still
# conveyed, just without colour. Levels: "good" (green), "warning" (amber),
# "danger" (red), "neutral" (grey, no tag).
.status_colours <- list(
  good    = list(bg = "#2E7D32", tag = "OK"),
  warning = list(bg = "#B36B00", tag = "FLAG"),
  danger  = list(bg = "#B00020", tag = "ALERT"),
  neutral = list(bg = "#616161", tag = NULL)
)

.status_badge <- function(text, level = "neutral") {
  spec <- .status_colours[[level]]
  if (is.null(spec)) spec <- .status_colours$neutral

  # Degrade gracefully to the plain-text tag (same as docx) if kableExtra
  # isn't installed, rather than erroring on this optional dependency.
  if (.is_docx || !requireNamespace("kableExtra", quietly = TRUE)) {
    if (is.null(spec$tag)) return(text)
    return(paste0("[", spec$tag, "] ", text))
  }

  fmt <- if (.is_pdf) "latex" else "html"
  kableExtra::cell_spec(text, fmt, color = "white", bold = TRUE,
                        background = spec$bg)
}

# -- Safe helpers --------------------------------------------------------------
# Self-contained: no dependency on %||% or rlang, which are absent from
# quarto::quarto_render()'s fresh environment when called from a Shiny app.

.null_or <- function(x, default) {
  if (is.null(x) || length(x) == 0L) default else x
}

.safe_round <- function(x, digits = 4) {
  x <- suppressWarnings(as.numeric(x))
  round(x, digits)
}

.safe_chr <- function(x, fallback = "N/A") {
  if (is.null(x) || length(x) == 0L || (length(x) == 1L && is.na(x)))
    return(fallback)
  as.character(x[[1L]])
}

.fs <- function(s, nm) {
  v <- s[[nm]]
  if (is.null(v) || length(v) == 0L) NA_real_
  else suppressWarnings(as.numeric(v[[1L]]))
}

# CrI with fallback for mixture priors that lack q025/q975.
# Mixture priors store only mean and sd; approximate with mean +/- 1.96 * sd.
.cri <- function(s) {
  lo <- .fs(s, "q025"); hi <- .fs(s, "q975")
  mn <- .fs(s, "mean"); sd_v <- .fs(s, "sd")
  if (is.na(lo)) lo <- mn - 1.96 * sd_v
  if (is.na(hi)) hi <- mn + 1.96 * sd_v
  paste0("[", .safe_round(lo, 4), ",  ", .safe_round(hi, 4), "]")
}

# For mixture priors, pull method/expert from the dominant component so
# these fields never display as "N/A" in tables.
.prior_method <- function(p) {
  if (!identical(p$dist, "mixture")) return(.safe_chr(p$method))
  dom <- p$components[[which.max(p$weights)]]
  paste0(.safe_chr(dom$method), " (pooled)")
}

.prior_expert <- function(p) {
  if (!identical(p$dist, "mixture")) return(.safe_chr(p$expert_id))
  nms <- vapply(p$components,
                function(x) .safe_chr(x$expert_id), character(1L))
  paste(unique(nms[nms != "N/A"]), collapse = ", ")
}

.params_df <- function(prior_obj) {
  p <- prior_obj$params
  if (is.null(p) || length(p) == 0L)
    return(data.frame(Parameter = character(0L), Value = character(0L),
                      stringsAsFactors = FALSE))
  if (identical(prior_obj$dist, "mixture")) {
    w <- prior_obj$weights
    if (is.null(w)) w <- p$weights
    if (is.null(w)) w <- numeric(0L)
    return(data.frame(
      Parameter        = paste0("weight[", seq_along(w), "]"),
      Value            = as.character(round(w, 6L)),
      stringsAsFactors = FALSE
    ))
  }
  vals <- tryCatch(unlist(p), error = function(e) NULL)
  if (is.null(vals) || !is.numeric(vals))
    return(data.frame(Parameter = character(0L), Value = character(0L),
                      stringsAsFactors = FALSE))
  data.frame(
    Parameter        = names(vals),
    Value            = as.character(.safe_round(vals, 6L)),
    stringsAsFactors = FALSE
  )
}

# Suppresses the expected mixture-family density warning inside plot calls.
.safe_plot <- function(expr_fn, msg = "Plot could not be rendered") {
  withCallingHandlers(
    tryCatch(expr_fn(), error = function(e) cat(msg, ":", conditionMessage(e), "\n")),
    warning = function(w) {
      if (grepl("different distribution families", conditionMessage(w), fixed = TRUE))
        invokeRestart("muffleWarning")
    }
  )
}
```

---

## Executive Summary

This report documents the Bayesian prior specification for study
**`r .safe_chr(params$trial_name, "Not specified")`**, prepared by
**`r .safe_chr(params$author, "Not specified")`**
(`r .safe_chr(params$sponsor, "Not specified")`) and dated `r format(Sys.Date(), "%d %B %Y")`.

```{r exec-table}
if (has_prior) {
  s <- pr$fit_summary
  df_exec <- data.frame(
    Item = c("Distribution family", "Elicitation method",
             "Expert / source", "Prior mean", "Prior SD",
             "95% Credible interval"),
    Value = c(
      toupper(.safe_chr(pr$dist)),
      .prior_method(pr),
      .prior_expert(pr),
      as.character(.safe_round(.fs(s, "mean"), 4)),
      as.character(.safe_round(.fs(s, "sd"),   4)),
      .cri(s)
    ),
    stringsAsFactors = FALSE
  )
  rownames(df_exec) <- NULL
  kable(df_exec, col.names = c("Item", "Value"), align = "ll",
        caption = "Prior specification summary", row.names = FALSE)
}
```

```{r key-findings}
#| results: asis
findings <- character(0L)

if (has_conf) {
  sev <- tolower(.safe_chr(cd$conflict_severity, "none"))
  bp  <- .safe_round(cd$box_pvalue, 3)
  if (sev == "none") {
    findings <- c(findings, paste0(
      "No prior-data conflict detected (Box p = ", bp, ")."))
  } else {
    findings <- c(findings, paste0(
      "**Prior-data conflict flagged** (severity: ", toupper(sev),
      ", Box p = ", bp, ") -- see Section 4."))
  }
}

if (has_sens) {
  sc <- sa$influence_scores
  any_sensitive <- any(is.na(sc) | sc > 0.05)
  if (any_sensitive) {
    worst <- names(sc)[which.max(ifelse(is.na(sc), Inf, sc))]
    findings <- c(findings, paste0(
      "**Posterior conclusions are sensitive to prior specification** ",
      "(", .safe_chr(worst, "at least one target"),
      " exceeds the sensitivity threshold) -- see Section 5."))
  } else {
    findings <- c(findings,
      "Posterior conclusions are robust to plausible variation in the prior's hyperparameters.")
  }
}

if (has_sens && !has_robust) {
  any_sensitive <- has_sens && any(is.na(sa$influence_scores) | sa$influence_scores > 0.05)
  if (any_sensitive) {
    findings <- c(findings, paste0(
      "No robust, sceptical, or power prior comparison was performed -- ",
      "given the sensitivity flagged above, one is recommended before ",
      "relying on this prior specification. See Section 6."))
  }
} else if (!has_robust) {
  findings <- c(findings,
    "No robust, sceptical, or power prior comparison was performed -- see Section 6.")
}

if (length(findings) > 0L) {
  .callout("important", paste(
    paste0("- ", findings), collapse = "\n"
  ))
}
```

```{r notes-block}
#| results: asis
note_text <- .null_or(params$notes, "")
if (nzchar(note_text)) {
  .callout("note", paste0("**Statistician's notes:** ", note_text))
}
```

---

## Trial Information

```{r trial-info}
df_trial <- data.frame(
  Field = c("Trial / Protocol", "Sponsor", "Statistician", "Report date"),
  Value = c(
    .safe_chr(params$trial_name, "Not specified"),
    .safe_chr(params$sponsor, "Not specified"),
    .safe_chr(params$author, "Not specified"),
    format(Sys.Date(), "%d %B %Y")
  ),
  stringsAsFactors = FALSE
)
rownames(df_trial) <- NULL
kable(df_trial, col.names = c("Field", "Value"), align = "ll",
      row.names = FALSE)
```

---

## Prior Specification

### Elicitation Methodology

```{r elicitation-method}
if (has_prior) {
  df_elic <- data.frame(
    Attribute = c("Quantity", "Distribution", "Method", "Expert / Source"),
    Value     = c(
      .safe_chr(pr$label),
      toupper(.safe_chr(pr$dist)),
      .prior_method(pr),
      .prior_expert(pr)
    ),
    stringsAsFactors = FALSE
  )
  rownames(df_elic) <- NULL
  kable(df_elic, col.names = c("Attribute", "Value"), align = "ll",
        row.names = FALSE)
}
```

### Fitted Parameters

```{r fitted-params}
#| results: asis
if (has_prior) {
  if (identical(pr$dist, "mixture")) {
    # For mixture priors: show component summary table instead of raw params
    n_comp <- length(pr$components)
    df_mix <- do.call(rbind, lapply(seq_len(n_comp), function(i) {
      comp <- pr$components[[i]]
      s    <- comp$fit_summary
      data.frame(
        Component  = paste0("Component ", i,
                            " (", toupper(.safe_chr(comp$dist)), ")"),
        Weight     = as.character(round(pr$weights[[i]], 4)),
        Expert     = .safe_chr(comp$expert_id),
        Mean       = as.character(.safe_round(.fs(s, "mean"), 4)),
        SD         = as.character(.safe_round(.fs(s, "sd"), 4)),
        stringsAsFactors = FALSE
      )
    }))
    rownames(df_mix) <- NULL
    kable(df_mix,
          col.names = c("Component", "Weight", "Expert", "Mean", "SD"),
          caption   = "Mixture prior: component summary",
          align     = "lllrr", row.names = FALSE)
  } else {
    df <- .params_df(pr)
    rownames(df) <- NULL
    if (nrow(df) > 0L) {
      kable(df, col.names = c("Parameter", "Value"),
            caption = "Fitted hyperparameters", align = "lr",
            row.names = FALSE)
    }
  }
}
```

### Prior Summary Statistics

```{r prior-summary}
if (has_prior) {
  s    <- pr$fit_summary
  mn   <- .fs(s, "mean"); sd_v <- .fs(s, "sd")
  lo   <- .fs(s, "q025"); if (is.na(lo)) lo <- mn - 1.96 * sd_v
  md   <- .fs(s, "q500"); if (is.na(md)) md <- mn
  hi   <- .fs(s, "q975"); if (is.na(hi)) hi <- mn + 1.96 * sd_v
  df_sum <- data.frame(
    Statistic = c("Mean", "Standard deviation",
                  "2.5th percentile", "Median", "97.5th percentile"),
    Value     = as.character(.safe_round(c(mn, sd_v, lo, md, hi), 5)),
    stringsAsFactors = FALSE
  )
  rownames(df_sum) <- NULL
  kable(df_sum, caption = "Prior distribution summary statistics",
        align = "lr", row.names = FALSE)
}
```

### Prior Density

```{r prior-density}
#| fig-cap: "Elicited prior distribution. Shaded region = 95% credible interval; dashed line = prior mean."
if (has_prior) {
  if (!is.null(pp_prior_plot) && inherits(pp_prior_plot, "gg")) {
    print(pp_prior_plot)
  } else {
    .safe_plot(function() plot(pr), "Prior density plot could not be rendered")
  }
}
```

---

## Prior-Data Conflict Assessment

```{r conflict-check}
#| results: asis
if (!has_conf) {
  .callout("warning", "No prior-data conflict diagnostics were computed for this report.")
}
```

```{r conflict-table}
if (has_conf) {
  df_conf <- data.frame(
    Diagnostic = c(
      "Box's prior predictive p-value",
      "Surprise index (standardised distance)",
      "KL divergence (prior || likelihood)",
      "Bhattacharyya overlap coefficient",
      "Conflict severity"
    ),
    Value = c(
      as.character(.safe_round(cd$box_pvalue,          4)),
      as.character(.safe_round(cd$surprise_index,      4)),
      as.character(.safe_round(cd$kl_prior_likelihood, 4)),
      as.character(.safe_round(cd$overlap,             4)),
      {
        sev_val   <- tolower(.safe_chr(cd$conflict_severity, "none"))
        sev_level <- switch(sev_val, none = "good", mild = "warning",
                            severe = "danger", "neutral")
        .status_badge(toupper(.safe_chr(cd$conflict_severity)), sev_level)
      }
    ),
    Interpretation = c(
      ifelse(isTRUE(cd$conflict_flag),
             paste0("Conflict detected (p < alpha = ", cd$alpha, ")"),
             "No conflict detected"),
      {
        si <- suppressWarnings(as.numeric(cd$surprise_index))
        if (is.na(si)) "Could not compute"
        else if (si > 3) "High surprise (> 3 SD)"
        else if (si > 2) "Moderate surprise"
        else "Within expected range"
      },
      {
        kl <- suppressWarnings(as.numeric(cd$kl_prior_likelihood))
        if (is.na(kl)) "Could not compute"
        else if (kl > 1) "Large divergence"
        else "Acceptable divergence"
      },
      {
        ov <- suppressWarnings(as.numeric(cd$overlap))
        if (is.na(ov)) "Could not compute"
        else if (ov < 0.3) "Low overlap - concern"
        else if (ov < 0.6) "Moderate overlap"
        else "Good overlap"
      },
      {
        sev <- tolower(.safe_chr(cd$conflict_severity))
        if (sev == "none") "No corrective action needed"
        else if (sev == "mild") "Consider reviewing the prior specification"
        else if (sev == "severe") "Prior specification should be reconsidered"
        else "Could not compute"
      }
    ),
    stringsAsFactors = FALSE
  )
  rownames(df_conf) <- NULL
  kable(df_conf, caption = "Prior-data conflict diagnostics",
        align = "llr", row.names = FALSE, escape = FALSE)
}
```

```{r conflict-recommendation}
#| results: asis
if (has_conf) {
  sev  <- .safe_chr(cd$conflict_severity, fallback = "none")
  .callout(
    switch(.safe_chr(cd$conflict_severity, fallback="none"),
      none="tip", mild="warning", severe="important", "note"),
    .safe_chr(cd$recommendation)
  )
}
```

### Prior - Likelihood - Posterior Overlay

```{r overlay-plot}
#| fig-cap: "Overlay of prior density, scaled likelihood, and resulting posterior."
if (has_conf && has_prior) {
  if (!is.null(pp_overlay_plot) && inherits(pp_overlay_plot, "gg")) {
    print(pp_overlay_plot)
  } else {
    .safe_plot(
      function() plot_prior_likelihood(pr, cd$data_summary, show_posterior = TRUE),
      "Overlay plot could not be rendered"
    )
  }
}
```

---

## Sensitivity Analysis

```{r sensitivity-check}
#| results: asis
if (!has_sens) {
  .callout("warning", "No sensitivity analysis was conducted for this report.")
}
```

### Influence of Prior Hyperparameters

```{r sensitivity-influence}
if (has_sens) {
  sc  <- sa$influence_scores
  nms <- names(sc)
  if (is.null(nms) || length(nms) == 0L)
    nms <- paste0("target_", seq_along(sc))
  is_sensitive <- is.na(sc) | sc > 0.05
  df_sens <- data.frame(
    Target    = nms,
    Range     = as.character(.safe_round(sc, 5)),
    Sensitive = mapply(function(txt, flag)
      .status_badge(txt, if (flag) "warning" else "good"),
      ifelse(is_sensitive, "Yes", "No"), is_sensitive),
    stringsAsFactors = FALSE
  )
  rownames(df_sens) <- NULL
  kable(df_sens,
        caption = "Posterior sensitivity to prior hyperparameters",
        align = "lrr", row.names = FALSE, escape = FALSE)
}
```

### Tornado Plot

```{r tornado}
#| fig-cap: "Range of each posterior quantity across the hyperparameter grid. Wider bars = greater prior influence."
if (has_sens) {
  if (!is.null(pp_tornado_plot) && inherits(pp_tornado_plot, "gg")) {
    print(pp_tornado_plot)
  } else {
    .safe_plot(function() plot_tornado(sa), "Tornado plot could not be rendered")
  }
}
```

### Hyperparameter Influence Heatmap

```{r heatmap}
#| fig-cap: "Posterior mean across the two-dimensional hyperparameter grid."
if (has_sens) {
  if (!is.null(pp_heatmap_plot) && inherits(pp_heatmap_plot, "gg")) {
    print(pp_heatmap_plot)
  } else {
    .safe_plot(
      function() plot_sensitivity(sa, target = sa$target[[1L]]),
      "Heatmap could not be rendered"
    )
  }
}
```

---

## Robust and Sensitivity Priors

```{r robust-check}
#| results: asis
if (!has_robust) {
  sens_flagged <- has_sens &&
    any(is.na(sa$influence_scores) | sa$influence_scores > 0.05)
  msg <- if (sens_flagged) {
    paste0(
      "No robust, sceptical, or power priors were computed for this ",
      "report. Section 5 flagged sensitivity to the prior specification ",
      "for at least one posterior target -- a robust or sceptical prior ",
      "comparison is recommended before relying on this specification."
    )
  } else {
    "No robust, sceptical, or power priors were computed for this report."
  }
  .callout("warning", msg)
}
```

```{r robust-mixture}
#| fig-cap: "Robust mixture prior: informative component mixed with a vague component."
if (!is.null(rob_prior)) {
  s <- rob_prior$fit_summary
  df_rob <- data.frame(
    Attribute = c("Type", "Vague weight", "Informative weight",
                  "Mixture mean", "Mixture SD"),
    Value     = c("Robust mixture",
                  paste0(round(rob_prior$vague_weight * 100), "%"),
                  paste0(round((1 - rob_prior$vague_weight) * 100), "%"),
                  as.character(.safe_round(.fs(s, "mean"), 4)),
                  as.character(.safe_round(.fs(s, "sd"),   4))),
    stringsAsFactors = FALSE
  )
  rownames(df_rob) <- NULL
  kable(df_rob, col.names = c("Attribute", "Value"), align = "ll",
        caption = "Robust mixture prior summary", row.names = FALSE)
  if (!is.null(pp_rob_plot) && inherits(pp_rob_plot, "gg")) {
    print(pp_rob_plot)
  } else {
    .safe_plot(function() plot(rob_prior), "Robust prior plot could not be rendered")
  }
}
```

```{r sceptical-prior}
#| fig-cap: "Sceptical prior centred at the null treatment effect."
if (!is.null(scep_prior)) {
  s <- scep_prior$fit_summary
  df_scep <- data.frame(
    Attribute = c("Type", "Family", "Null value", "Mean", "SD"),
    Value     = c("Sceptical",
                  toupper(.safe_chr(scep_prior$dist)),
                  .safe_chr(if(is.null(scep_prior$sceptical_null)) NULL else scep_prior$sceptical_null),
                  as.character(.safe_round(.fs(s, "mean"), 4)),
                  as.character(.safe_round(.fs(s, "sd"),   4))),
    stringsAsFactors = FALSE
  )
  rownames(df_scep) <- NULL
  kable(df_scep, col.names = c("Attribute", "Value"), align = "ll",
        caption = "Sceptical prior summary", row.names = FALSE)
  if (!is.null(pp_scep_plot) && inherits(pp_scep_plot, "gg")) {
    print(pp_scep_plot)
  } else {
    .safe_plot(function() plot(scep_prior), "Sceptical prior plot could not be rendered")
  }
}
```

```{r power-prior}
#| fig-cap: "Power prior calibration curves."
if (!is.null(pow_result)) {
  df_pow <- data.frame(
    Attribute = c("Method", "Target BF", "Optimal delta",
                  "Power prior mean", "Power prior SD"),
    Value     = c(.safe_chr(pow_result$method),
                  as.character(if(is.null(pow_result$target_bf)) "-" else pow_result$target_bf),
                  as.character(pow_result$delta_opt),
                  as.character(.safe_round(
                    .fs(pow_result$power_prior$fit_summary, "mean"), 4)),
                  as.character(.safe_round(
                    .fs(pow_result$power_prior$fit_summary, "sd"), 4))),
    stringsAsFactors = FALSE
  )
  rownames(df_pow) <- NULL
  kable(df_pow, col.names = c("Attribute", "Value"), align = "ll",
        caption = "Power prior calibration summary", row.names = FALSE)
  if (!is.null(pp_pow_plot) && inherits(pp_pow_plot, "gg")) {
    print(pp_pow_plot)
  } else {
    .safe_plot(function() plot(pow_result), "Power prior plot could not be rendered")
  }
}
```

---

## Regulatory Compliance Checklist

```{r compliance}
df_comp <- data.frame(
  Requirement = c(
    "Prior elicitation method documented",
    "Distribution family and parameters specified",
    "Expert / source of prior identified",
    "Prior density plot provided",
    "Prior-data conflict assessed",
    "Conflict diagnostic statistics reported",
    "Sensitivity analysis performed",
    "Sensitivity visualisations provided",
    "Alternative priors considered",
    "Robust / sceptical prior computed",
    "Regulatory report generated"
  ),
  Status = c(
    ifelse(has_prior,  "Complete",  "Missing"),
    ifelse(has_prior,  "Complete",  "Missing"),
    ifelse(has_prior,  "Complete",  "Missing"),
    ifelse(has_prior,  "Complete",  "Missing"),
    ifelse(has_conf,   "Complete",  "Not run"),
    ifelse(has_conf,   "Complete",  "Not run"),
    ifelse(has_sens,   "Complete",  "Not run"),
    ifelse(has_sens,   "Complete",  "Not run"),
    ifelse(has_sens,   "See above", "Not run"),
    ifelse(has_robust, "Complete",  "Not run"),
    "Complete"
  ),
  stringsAsFactors = FALSE
)
# "Missing" (a required prior was never supplied) is a more serious gap
# than "Not run" (an optional downstream analysis wasn't performed).
.status_level_map <- c(Complete = "good", Missing = "danger",
                       "Not run" = "warning", "See above" = "neutral")
df_comp$Status <- mapply(function(txt) .status_badge(txt, .status_level_map[[txt]]),
                         df_comp$Status)
rownames(df_comp) <- NULL
kable(df_comp, align = "ll",
      caption = "FDA/EMA regulatory compliance checklist",
      row.names = FALSE, escape = FALSE)
```

```{r fda-note, results="asis"}
.callout("note", paste0(
  "This checklist maps to considerations in the FDA's 2026 draft guidance, ",
  "Use of Bayesian Methodology in Clinical Trials of Drug and Biological ",
  "Products, and EMA's 2026 concept paper on the development of a reflection ",
  "paper on the use of Bayesian methods in clinical development ",
  "(EMA/CHMP/1813/2026)."
))
```

---

## Appendix: Session Information

```{r session-info}
si <- sessionInfo()
df_sess <- data.frame(
  Item  = c("R version",
            "bayprior version",
            "quarto R package",
            "Quarto CLI version",
            "Platform",
            "Date"),
  Value = c(
    paste(si$R.version$major, si$R.version$minor, sep = "."),
    tryCatch(as.character(utils::packageVersion("bayprior")),
             error = function(e) "unknown"),
    tryCatch(as.character(utils::packageVersion("quarto")),
             error = function(e) "unknown"),
    tryCatch(as.character(quarto::quarto_version()),
             error = function(e) "not found - check Quarto CLI installation"),
    si$platform,
    format(Sys.Date(), "%d %B %Y")
  ),
  stringsAsFactors = FALSE
)
rownames(df_sess) <- NULL
kable(df_sess, align = "ll", row.names = FALSE)
```

---

*Report generated by the **bayprior** R package using `prior_report()`.
Methods follow O'Hagan et al. (2006), Box (1980), Schmidli et al. (2014),
Ibrahim & Chen (2000), and the FDA's 2026 draft guidance, Use of Bayesian
Methodology in Clinical Trials of Drug and Biological Products.*
