---
title: "Introduction to bayprior"
author: "bayprior package"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to bayprior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 6,
  fig.height = 3.8,
  out.width  = "95%",
  dpi        = 120,
  warning    = FALSE,
  message    = FALSE
)
```

# Overview

**bayprior** is an R package for principled Bayesian prior elicitation,
conflict detection, and sensitivity analysis in clinical trials. It implements
the structured methodology recommended in the FDA Draft Guidance on Bayesian
Statistical Methods (2026) and the SHELF elicitation framework
(O'Hagan et al., 2006).

The core analytical workflow is:

1. **Elicit** — Specify a prior from expert knowledge or historical data
2. **Pool** — Combine priors from multiple experts into a consensus prior
3. **Diagnose** — Detect conflict between the prior and observed data
4. **Analyse** — Assess how sensitive posterior conclusions are to prior choice
5. **Robust** — Construct robust, sceptical, or power priors as alternatives
6. **Report** — Generate a regulatory-ready prior justification report

---

# Prior Elicitation

## Beta Prior (Binary Endpoints)

The Beta distribution is the natural conjugate prior for a binary response rate
$\theta \in (0, 1)$. `elicit_beta()` supports two elicitation methods.

### Quantile Matching

The expert specifies quantiles directly — for example: "I believe the true
response rate has a 5% chance of being below 10%, a median of around 30%, and
a 95% chance of being below 60%."

```{r beta-quantile}
library(bayprior)

prior_q <- elicit_beta(
  quantiles = c("0.05" = 0.10, "0.50" = 0.30, "0.95" = 0.60),
  expert_id = "Expert_1",
  label     = "Response rate (treatment arm)"
)
print(prior_q)
```

### Moment Matching

Alternatively, elicit the prior mean and standard deviation directly:

```{r beta-moments}
prior_m <- elicit_beta(
  mean      = 0.30,
  sd        = 0.10,
  method    = "moments",
  expert_id = "Expert_1",
  label     = "Response rate (treatment arm)"
)
plot(prior_m)
```

The shaded region shows the 95% credible interval and the dashed vertical line
marks the prior mean.

## Normal Prior (Continuous Endpoints)

For continuous quantities such as mean differences or log odds ratios:

```{r normal-prior}
prior_nor <- elicit_normal(
  quantiles = c("0.025" = -0.5, "0.50" = 0.20, "0.975" = 0.90),
  label     = "Log odds ratio"
)
print(prior_nor)
```

## Gamma Prior (Rate / Count Endpoints)

For positive-valued quantities such as Poisson event rates or median survival:

```{r gamma-prior}
prior_gam <- elicit_gamma(
  mean   = 5,
  sd     = 2,
  method = "moments",
  label  = "Median OS (months)"
)
plot(prior_gam)
```

## Log-Normal Prior (Hazard Ratios)

For multiplicative quantities that must remain positive:

```{r lognormal-prior}
prior_ln <- elicit_lognormal(
  quantiles = c("0.05" = 0.40, "0.50" = 0.70, "0.95" = 1.20),
  label     = "Hazard ratio (treatment vs control)"
)
plot(prior_ln)
```

## Roulette Method

The roulette method (Oakley & O'Hagan, 2010) converts chip allocations across
pre-defined bins into a parametric prior via quantile matching. It is available
interactively through the Shiny app and programmatically here:

```{r roulette}
prior_rou <- elicit_roulette(
  chips  = c(0L, 2L, 5L, 8L, 5L, 2L, 1L),
  breaks = seq(0, 0.7, by = 0.1),
  family = "beta",
  label  = "Response rate"
)
print(prior_rou)
```

---

# Expert Pooling

When multiple experts contribute, their priors are aggregated into a single
consensus prior via `aggregate_experts()`.

## Linear Pooling

Linear pooling produces a weighted mixture of the individual priors. It is the
most common approach in clinical trial settings (externally Bayesian):

```{r linear-pooling}
p1 <- elicit_beta(mean = 0.25, sd = 0.08, method = "moments",
                  expert_id = "E1", label = "Response rate")
p2 <- elicit_beta(mean = 0.35, sd = 0.10, method = "moments",
                  expert_id = "E2", label = "Response rate")
p3 <- elicit_beta(mean = 0.30, sd = 0.09, method = "moments",
                  expert_id = "E3", label = "Response rate")

consensus <- aggregate_experts(
  priors  = list(E1 = p1, E2 = p2, E3 = p3),
  weights = c(0.40, 0.30, 0.30),
  method  = "linear"
)
print(consensus)
plot(consensus)
```

`aggregate_experts()` automatically computes pairwise Bhattacharyya
coefficients and warns when expert disagreement is substantial.

## Manual Mixture Construction

You can also build a mixture prior directly from components:

```{r manual-mixture}
mix <- elicit_mixture(
  components = list(p1, p2),
  weights    = c(0.5, 0.5),
  label      = "50-50 pooled prior"
)
plot(mix)
```

---

# Prior-Data Conflict Diagnostics

Before updating the prior with trial data it is good practice to check
compatibility between the prior and the observed data. `prior_conflict()`
computes four complementary diagnostics:

| Diagnostic | Interpretation |
|---|---|
| Box p-value | Prior predictive p-value; < 0.05 flags conflict |
| Surprise index | Standardised distance; > 3 indicates high surprise |
| KL divergence | Information distance from prior to likelihood |
| Bhattacharyya overlap | Distributional overlap; < 0.3 is concerning |

```{r conflict}
prior <- elicit_beta(
  mean   = 0.30,
  sd     = 0.10,
  method = "moments",
  label  = "Response rate"
)

# Observed: 18 events in 40 patients
cd <- prior_conflict(
  prior        = prior,
  data_summary = list(type = "binary", x = 18, n = 40),
  alpha        = 0.05
)
print(cd)
```

Visualise the prior-likelihood-posterior overlay to assess compatibility
graphically:

```{r overlay, fig.height = 4}
plot_prior_likelihood(
  prior,
  data_summary   = list(type = "binary", x = 18, n = 40),
  show_posterior = TRUE
)
```

When the prior (blue) and likelihood (orange) overlap substantially there is no
meaningful conflict, and the posterior (green) represents a credible update.

---

# Sensitivity Analysis

Regulatory guidelines require demonstrating that trial conclusions are robust
to plausible prior variations. `sensitivity_grid()` evaluates how posterior
inferences change across a grid of hyperparameter values.

Note: the grids below are intentionally coarse for vignette build speed. In
practice, use finer grids (e.g. `seq(1, 8, 0.5)`) to obtain smooth surfaces.

```{r sensitivity, cache = TRUE}
sa <- sensitivity_grid(
  prior        = prior,
  data_summary = list(type = "binary", x = 14, n = 40),
  param_grid   = list(alpha = seq(1, 6, 1), beta = seq(2, 14, 2)),
  target       = c("posterior_mean", "prob_efficacy"),
  threshold    = 0.30
)
```

## Tornado Plot

The tornado plot ranks posterior quantities by their range across the
hyperparameter grid. Wider bars indicate higher prior sensitivity:

```{r tornado, fig.height = 3}
plot_tornado(sa)
```

## Influence Heatmap

The heatmap shows how the posterior mean changes as both hyperparameters vary.
The orange diamond marks the reference prior:

```{r heatmap, fig.height = 4}
plot_sensitivity(sa, target = "posterior_mean")
```

## Credible Interval Sensitivity

`sensitivity_cri()` focuses specifically on the posterior credible interval
width — a key regulatory quantity:

```{r cri-sensitivity, cache = TRUE, fig.height = 4}
cri_sa <- sensitivity_cri(
  prior        = prior,
  data_summary = list(type = "binary", x = 14, n = 40),
  param_grid   = list(alpha = seq(1, 6, 1), beta = seq(2, 14, 2)),
  cri_level    = 0.95,
  threshold    = 0.30
)
plot_sensitivity(cri_sa, target = "cri_width")
```

---

# Robust and Sceptical Priors

When prior-data conflict is detected, or when a conservative regulatory stance
is required, bayprior provides ready-made alternative prior constructions.

## Robust Mixture Prior

A robust prior mixes an informative component with a vague (diffuse) component
(Schmidli et al., 2014). The vague component ensures the posterior is never
dominated by a conflicting informative prior. The default `vague_weight = 0.20`
gives an 80/20 informative/vague split:

```{r robust, fig.height = 3.5}
informative <- elicit_beta(
  mean   = 0.30,
  sd     = 0.08,
  method = "moments",
  label  = "Response rate"
)

rob <- robust_prior(
  informative  = informative,
  vague_weight = 0.20,
  label        = "Robust mixture prior"
)
plot(rob)
```

## Sceptical Prior

A sceptical prior is centred at the null value of the treatment effect and
represents a conservative regulatory stance (Spiegelhalter et al., 1994).

For a **Normal** family (e.g. log odds ratio, mean difference), `null_value`
is the null treatment effect (typically 0):

```{r sceptical-normal, fig.height = 3.5}
sc_norm <- sceptical_prior(
  null_value = 0,
  family     = "normal",
  strength   = "moderate",
  label      = "Log odds ratio (sceptical)"
)
plot(sc_norm)
```

For a **Beta** family (binary response rate), `null_value` must be in (0, 1):

```{r sceptical-beta, fig.height = 3.5}
sc_beta <- sceptical_prior(
  null_value = 0.20,   # null response rate of 20%
  family     = "beta",
  strength   = "moderate",
  label      = "Response rate (sceptical)"
)
plot(sc_beta)
```

---

# Power Prior Calibration

When relevant historical data exist, a power prior down-weights the historical
evidence by $\delta \in (0, 1]$ before incorporating it. `calibrate_power_prior()`
selects $\delta$ to achieve a target Bayes Factor (Ibrahim & Chen, 2000;
Gravestock & Held, 2017).

```{r power-prior, cache = TRUE}
base <- elicit_beta(
  mean   = 0.50,
  sd     = 0.20,
  method = "moments",
  label  = "Response rate"
)

calib <- calibrate_power_prior(
  historical_data = list(type = "binary", x = 12, n = 40),
  current_data    = list(type = "binary", x = 18, n = 50),
  base_prior      = base,
  target_bf       = 3,
  delta_grid      = seq(0.10, 1.0, by = 0.10),  # coarse grid for vignette speed
  method          = "bayes_factor"
)
print(calib)
plot(calib)
```

The calibration curves show the Bayes Factor (top) and Box p-value (bottom)
across the $\delta$ grid. The vertical dotted green line marks the optimal
weight and the horizontal dashed lines mark the target BF and $\alpha = 0.05$.

---

# Multivariate Conflict Diagnostics

For trials with co-primary endpoints, `conflict_mahalanobis()` extends the
conflict check to the multivariate setting. The Mahalanobis distance has a
known $\chi^2$ reference distribution under compatibility:

```{r mahal}
mv <- conflict_mahalanobis(
  prior_means = c(0.35, 0.60),
  prior_cov   = matrix(c(0.01, 0.003, 0.003, 0.015), 2, 2),
  obs_means   = c(0.55, 0.58),
  obs_cov     = matrix(c(0.008, 0.002, 0.002, 0.010), 2, 2) / 50,
  labels      = c("Response rate", "OS rate")
)
print(mv)
```

The `marginal_z` scores identify which endpoint contributes most to any
detected conflict.

---

# Generating a Prior Justification Report

Once the analysis is complete, `prior_report()` renders a self-contained
regulatory-ready document in HTML, PDF, or Word format. The report includes
the prior specification, conflict diagnostics, sensitivity visualisations, and
an FDA/EMA compliance checklist.

```{r report, eval = FALSE}
# prior_report() calls rmarkdown::render() internally and cannot be run
# inside a vignette build. Run interactively after loading bayprior.

prior_report(
  prior         = prior,
  conflict      = cd,
  sensitivity   = sa,
  output_format = "html",       # or "pdf" or "docx"
  trial_name    = "TRIAL-001",
  sponsor       = "BioPharma Ltd",
  author        = "J. Smith",
  notes         = paste0(
    "Prior based on Phase 2 internal data and two external expert ",
    "elicitations. Sensitivity analysis confirms robustness across a ",
    "wide range of prior hyperparameter values."
  )
)
```

---

# Complete Worked Example

The following end-to-end example mirrors a realistic binary-endpoint trial
workflow, combining three expert opinions and assessing sensitivity.

```{r full-example, cache = TRUE, fig.height = 4}
# 1. Elicit priors from three experts using different methods
e1 <- elicit_beta(mean = 0.28, sd = 0.08, method = "moments",
                  expert_id = "E1", label = "ORR")
e2 <- elicit_beta(mean = 0.35, sd = 0.10, method = "moments",
                  expert_id = "E2", label = "ORR")
e3 <- elicit_beta(
  quantiles = c("0.10" = 0.15, "0.50" = 0.30, "0.90" = 0.52),
  expert_id = "E3", label = "ORR"
)

# 2. Pool into a consensus prior
consensus <- aggregate_experts(
  priors  = list(E1 = e1, E2 = e2, E3 = e3),
  weights = c(0.40, 0.35, 0.25),
  method  = "linear"
)

# 3. Check conflict with interim data (20 responses in 55 patients)
data_obs <- list(type = "binary", x = 20, n = 55)
cd_full  <- prior_conflict(consensus, data_obs)
print(cd_full)

# 4. Visualise prior-likelihood-posterior
plot_prior_likelihood(consensus, data_obs, show_posterior = TRUE)
```

```{r full-sensitivity, cache = TRUE, fig.height = 3}
# 5. Sensitivity analysis (coarse grid for vignette speed)
sa_full <- sensitivity_grid(
  prior        = consensus,
  data_summary = data_obs,
  param_grid   = list(alpha = seq(1, 6, 1), beta = seq(2, 12, 2)),
  target       = c("posterior_mean", "prob_efficacy"),
  threshold    = 0.25
)
plot_tornado(sa_full)
```

```{r full-heatmap, cache = TRUE, fig.height = 4}
plot_sensitivity(sa_full, target = "prob_efficacy")
```

```{r full-robust, fig.height = 3.5}
# 6. Robust alternative in case conflict worsens at final analysis
rob_full <- robust_prior(consensus, vague_weight = 0.20)
plot(rob_full)
```

---

# Function Reference

| Function | Purpose |
|---|---|
| `elicit_beta()` | Beta prior for binary endpoints |
| `elicit_normal()` | Normal prior for continuous endpoints |
| `elicit_gamma()` | Gamma prior for rate/count endpoints |
| `elicit_lognormal()` | Log-normal prior for hazard ratios |
| `elicit_roulette()` | Chip-based roulette elicitation |
| `elicit_mixture()` | Manual mixture prior construction |
| `aggregate_experts()` | Pool multiple expert priors |
| `prior_conflict()` | Univariate prior-data conflict diagnostics |
| `conflict_mahalanobis()` | Multivariate conflict diagnostics |
| `sensitivity_grid()` | Grid sensitivity of posterior quantities |
| `sensitivity_cri()` | Credible interval sensitivity |
| `plot_sensitivity()` | Heatmap / line plot of sensitivity results |
| `plot_tornado()` | Tornado plot of prior influence |
| `plot_prior_likelihood()` | Prior-likelihood-posterior overlay |
| `robust_prior()` | Robust mixture prior |
| `sceptical_prior()` | Sceptical prior centred at null |
| `calibrate_power_prior()` | Calibrated power prior from historical data |
| `prior_report()` | Regulatory prior justification report |

---

# References

Box, G. E. P. (1980). Sampling and Bayes' inference in scientific modelling and
robustness. *Journal of the Royal Statistical Society A*, 143, 383–430.

Gravestock, I. & Held, L. (2017). Adaptive power priors with empirical Bayes
for clinical trials. *Pharmaceutical Statistics*, 16, 349–360.

Ibrahim, J. G. & Chen, M.-H. (2000). Power prior distributions for regression
models. *Statistical Science*, 15, 46–60.

O'Hagan, A., Buck, C. E., Daneshkhah, A., Eiser, J. R., Garthwaite, P. H.,
Jenkinson, D. J., Oakley, J. E., & Rakow, T. (2006). *Uncertain Judgements:
Eliciting Experts' Probabilities*. Wiley.

Oakley, J. E. & O'Hagan, A. (2010). SHELF: the Sheffield Elicitation
Framework. University of Sheffield.

Schmidli, H., Gsteiger, S., Roychoudhury, S., O'Hagan, A., Spiegelhalter, D.,
& Neuenschwander, B. (2014). Robust meta-analytic-predictive priors in clinical
trials with historical control information. *Biometrics*, 70, 1023–1032.

Spiegelhalter, D. J., Freedman, L. S., & Parmar, M. K. B. (1994). Bayesian
approaches to randomized trials. *Journal of the Royal Statistical Society A*,
157, 357–416.
