---
title: "Prior-Data Conflict Diagnostics"
author: "bayprior package"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Prior-Data Conflict Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 6,
  fig.height = 3.8,
  out.width  = "95%",
  dpi        = 120,
  warning    = FALSE,
  message    = FALSE
)
library(bayprior)
```

# Why Conflict Diagnostics Matter

Before updating a prior with trial data it is essential to check whether the
two are compatible. Substantial prior-data conflict indicates one of three
things: the prior was mis-specified, the data are anomalous, or the model is
wrong. Regulators increasingly require evidence that conflict was assessed and
addressed.

**bayprior** provides four complementary univariate diagnostics and one
multivariate diagnostic, following Box (1980) and related work.

---

# Univariate Conflict: `prior_conflict()`

## Setting Up

```{r setup-prior}
prior <- elicit_beta(
  mean   = 0.30,
  sd     = 0.10,
  method = "moments",
  label  = "Response rate"
)
```

## No-Conflict Case

When prior and data are compatible:

```{r no-conflict}
cd_none <- prior_conflict(
  prior        = prior,
  data_summary = list(type = "binary", x = 13, n = 40),
  alpha        = 0.05
)
print(cd_none)
```

## Mild Conflict Case

When the observed data is somewhat surprising given the prior:

```{r mild-conflict}
cd_mild <- prior_conflict(
  prior        = prior,
  data_summary = list(type = "binary", x = 20, n = 40),
  alpha        = 0.05
)
print(cd_mild)
```

## Severe Conflict Case

When prior and data are strongly incompatible:

```{r severe-conflict}
cd_severe <- prior_conflict(
  prior        = prior,
  data_summary = list(type = "binary", x = 35, n = 40),
  alpha        = 0.05
)
print(cd_severe)
```

---

# The Four Diagnostics Explained

## 1. Box's Prior Predictive P-value

Proposed by Box (1980). Tests whether the observed data $y$ is plausible under
the prior predictive distribution $p(y) = \int p(y|\theta) \pi(\theta) d\theta$:

$$p_{\text{Box}} = P(|T(Y)| \geq |T(y_{\text{obs}})|)$$

where $T$ is a discrepancy measure. bayprior uses a Normal approximation:

$$z = \frac{\hat{\theta} - \mu_\pi}{\sqrt{\sigma_\pi^2 + \text{SE}^2}}, \quad
p_{\text{Box}} = 2\Phi(-|z|)$$

**Interpretation:** $p < 0.05$ flags conflict at the 5% level.

## 2. Surprise Index

The standardised distance between the prior mean and the observed data:

$$S = \left|\frac{\hat{\theta} - \mu_\pi}{\sqrt{\sigma_\pi^2 + \text{SE}^2}}\right|$$

**Interpretation:** $S > 2$ is moderate surprise; $S > 3$ is high surprise
(equivalent to a 3-sigma departure).

## 3. KL Divergence

The Kullback-Leibler divergence from the prior to the (normalised) likelihood,
approximated under normality:

$$\text{KL}(P \| Q) = \log\frac{\sigma_Q}{\sigma_P} +
\frac{\sigma_P^2 + (\mu_P - \mu_Q)^2}{2\sigma_Q^2} - \frac{1}{2}$$

**Interpretation:** Values > 1 indicate large information distance.

## 4. Bhattacharyya Overlap Coefficient

Quantifies distributional overlap between prior and likelihood:

$$\text{BC} = \exp\!\left(-\frac{1}{4}\log\frac{\sigma_P^2/\sigma_Q^2 +
\sigma_Q^2/\sigma_P^2 + 2}{4} - \frac{(\mu_P - \mu_Q)^2}{4(\sigma_P^2 + \sigma_Q^2)}\right)$$

**Interpretation:** 1 = identical distributions; 0 = no overlap. Values < 0.3
are concerning.

## Summary of Thresholds

```{r thresholds-table}
library(knitr)
kable(data.frame(
  Diagnostic       = c("Box p-value", "Surprise index",
                       "KL divergence", "Bhattacharyya overlap"),
  `No conflict`    = c(">= 0.05", "< 2", "< 0.5", "> 0.6"),
  `Mild conflict`  = c("0.01 - 0.05", "2 - 3", "0.5 - 1", "0.3 - 0.6"),
  `Severe conflict`= c("< 0.01", "> 3", "> 1", "< 0.3"),
  check.names      = FALSE
), align = "lrrr")
```

---

# Prior-Likelihood-Posterior Overlay

The canonical visualisation for conflict assessment:

```{r overlay-none, fig.cap="No conflict: prior and likelihood overlap substantially."}
plot_prior_likelihood(
  prior,
  data_summary   = list(type = "binary", x = 13, n = 40),
  show_posterior = TRUE
)
```

```{r overlay-severe, fig.cap="Severe conflict: prior (blue) and likelihood (orange) are widely separated."}
plot_prior_likelihood(
  prior,
  data_summary   = list(type = "binary", x = 35, n = 40),
  show_posterior = TRUE
)
```

When conflict is severe, the posterior (green) is pulled far from the prior,
and the prior has little influence on the conclusions.

---

# Continuous Endpoint Conflict

`prior_conflict()` handles continuous outcomes via the `type = "continuous"`
argument:

```{r continuous-conflict}
prior_norm <- elicit_normal(
  mean   = 0.0,
  sd     = 0.3,
  method = "moments",
  label  = "Log odds ratio"
)

cd_cont <- prior_conflict(
  prior        = prior_norm,
  data_summary = list(type = "continuous", x = 0.65, sd = 0.25, n = 80),
  alpha        = 0.05
)
print(cd_cont)
```

---

# Mixture Prior Conflict

Conflict diagnostics work transparently with mixture priors, using the mixture's
`fit_summary` (mean and SD) for the Normal approximation:

```{r mixture-conflict}
e1 <- elicit_beta(mean = 0.25, sd = 0.08, method = "moments",
                  expert_id = "E1", label = "Response rate")
e2 <- elicit_beta(mean = 0.40, sd = 0.10, method = "moments",
                  expert_id = "E2", label = "Response rate")

mix <- aggregate_experts(
  priors  = list(E1 = e1, E2 = e2),
  weights = c(0.5, 0.5)
)

cd_mix <- prior_conflict(
  prior        = mix,
  data_summary = list(type = "binary", x = 28, n = 50)
)
print(cd_mix)
```

---

# Multivariate Conflict: `conflict_mahalanobis()`

For trials with co-primary endpoints, the univariate diagnostics do not capture
the joint distribution. The Mahalanobis distance generalises Box's approach:

$$D^2 = (\hat{\boldsymbol{\theta}} - \boldsymbol{\mu}_\pi)^T
(\boldsymbol{\Sigma}_\pi + \boldsymbol{\Sigma}_{\text{lik}})^{-1}
(\hat{\boldsymbol{\theta}} - \boldsymbol{\mu}_\pi) \sim \chi^2_p$$

under compatibility, where $p$ is the number of endpoints.

```{r mahal}
mv <- conflict_mahalanobis(
  prior_means = c(0.35, 0.60),
  prior_cov   = matrix(c(0.010, 0.003, 0.003, 0.015), 2, 2),
  obs_means   = c(0.55, 0.58),
  obs_cov     = matrix(c(0.008, 0.002, 0.002, 0.010), 2, 2) / 50,
  labels      = c("Response rate", "OS rate"),
  alpha       = 0.05
)
print(mv)
```

The `marginal_z` scores identify which endpoint drives any detected conflict —
essential for guiding re-elicitation decisions.

---

# Conflict Severity Classification

bayprior automatically classifies severity and generates a plain-language
recommendation:

```{r severity}
cat(cd_none$recommendation, "\n\n")
cat(cd_mild$recommendation, "\n\n")
cat(cd_severe$recommendation, "\n")
```

---

# Regulatory Guidance

The FDA Draft Guidance on Bayesian Statistical Methods (2026) requires:

1. A pre-specified conflict checking procedure
2. Clear documentation of what action will be taken in each severity case
3. A sensitivity analysis demonstrating robustness to alternative priors

bayprior's conflict diagnostics directly support points 1 and 2. See the
**Sensitivity Analysis** vignette for point 3.

---

# References

Box, G. E. P. (1980). Sampling and Bayes' inference in scientific modelling and
robustness. *Journal of the Royal Statistical Society A*, 143, 383–430.

Spiegelhalter, D. J., Freedman, L. S., & Parmar, M. K. B. (1994). Bayesian
approaches to randomized trials. *Journal of the Royal Statistical Society A*,
157, 357–416.
