---
title: "Prior Elicitation Methods"
author: "bayprior package"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Prior Elicitation Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 6,
  fig.height = 3.8,
  out.width  = "95%",
  dpi        = 120,
  warning    = FALSE,
  message    = FALSE
)
library(bayprior)
```

# Overview

Prior elicitation is the process of translating an expert's subjective beliefs
about an unknown quantity into a formal probability distribution. **bayprior**
implements the structured elicitation framework recommended in the SHELF
methodology (O'Hagan et al., 2006) across four distribution families and three
elicitation methods.

## Distribution Families

| Family | Support | Typical quantity |
|---|---|---|
| **Beta** | (0, 1) | Response rates, proportions |
| **Normal** | (−∞, ∞) | Mean differences, log odds ratios |
| **Gamma** | (0, ∞) | Event rates, median survival |
| **Log-Normal** | (0, ∞) | Hazard ratios, PK parameters |

## Elicitation Methods

- **Quantile matching** — expert specifies probability–value pairs
- **Moment matching** — expert specifies mean and standard deviation
- **Roulette method** — expert allocates chips across histogram bins

---

# Beta Prior (Binary Endpoints)

## Quantile Matching

The expert specifies three quantiles that anchor the distribution. For example,
a clinician who believes the objective response rate (ORR) has a 5% chance of
being below 10%, a median of 30%, and a 95% chance of being below 60%:

```{r beta-quantile}
prior_q <- elicit_beta(
  quantiles = c("0.05" = 0.10, "0.50" = 0.30, "0.95" = 0.60),
  expert_id = "Expert_1",
  label     = "ORR (treatment arm)"
)
print(prior_q)
plot(prior_q)
```

The `quantiles` argument is a named numeric vector where names are
probabilities (as strings) and values are the corresponding quantiles.

## Moment Matching

When an expert finds it easier to specify a mean and standard deviation than
exact quantiles:

```{r beta-moments}
prior_m <- elicit_beta(
  mean      = 0.30,
  sd        = 0.10,
  method    = "moments",
  expert_id = "Expert_1",
  label     = "ORR (treatment arm)"
)
print(prior_m)
```

Moment matching solves analytically for the Beta parameters from the mean
and variance. The SD must be small enough that implied $\alpha, \beta > 0$:

$$\alpha = \mu \left(\frac{\mu(1-\mu)}{\sigma^2} - 1\right), \quad
\beta = (1-\mu)\left(\frac{\mu(1-\mu)}{\sigma^2} - 1\right)$$

## Choosing Between Methods

- **Use quantile matching** when the expert can more naturally think in terms of
  probability statements ("I'm 90% sure the rate is between X and Y")
- **Use moment matching** when the expert can directly state a best guess
  (mean) and uncertainty (SD)

---

# Normal Prior (Continuous Endpoints)

For quantities that can take any real value, such as log odds ratios or
mean differences:

```{r normal-quantile}
prior_lor <- elicit_normal(
  quantiles = c("0.025" = -0.50, "0.50" = 0.20, "0.975" = 0.90),
  label     = "Log odds ratio"
)
plot(prior_lor)
```

```{r normal-moments}
prior_md <- elicit_normal(
  mean      = 0.0,
  sd        = 0.5,
  method    = "moments",
  label     = "Mean difference (sceptical)"
)
plot(prior_md)
```

The Normal prior is conjugate for continuous outcomes with known variance,
making posterior updating straightforward.

---

# Gamma Prior (Rate and Count Endpoints)

For strictly positive quantities such as event rates or median survival times:

```{r gamma-moments}
prior_os <- elicit_gamma(
  mean   = 18,
  sd     = 6,
  method = "moments",
  label  = "Median OS (months)"
)
plot(prior_os)
```

```{r gamma-quantile}
prior_rate <- elicit_gamma(
  quantiles = c("0.10" = 2, "0.50" = 5, "0.90" = 10),
  label     = "Event rate (per 100 patient-years)"
)
print(prior_rate)
```

The Gamma parameterisation uses shape and rate:
$$\text{mean} = \frac{\text{shape}}{\text{rate}}, \quad
\text{SD} = \frac{\sqrt{\text{shape}}}{\text{rate}}$$

---

# Log-Normal Prior (Hazard Ratios)

For quantities that are inherently multiplicative and positive, such as hazard
ratios or pharmacokinetic parameters. Moment matching operates on the original
scale; parameters are stored on the log scale:

```{r lognormal-quantile}
prior_hr <- elicit_lognormal(
  quantiles = c("0.05" = 0.40, "0.50" = 0.70, "0.95" = 1.20),
  label     = "Hazard ratio (treatment vs control)"
)
print(prior_hr)
plot(prior_hr)
```

```{r lognormal-moments}
prior_pk <- elicit_lognormal(
  mean   = 25,
  sd     = 10,
  method = "moments",
  label  = "AUC (ng/mL*h)"
)
print(prior_pk)
```

The relationship between original-scale moments and log-scale parameters:
$$\sigma_{\log}^2 = \log\left(1 + \frac{\sigma^2}{\mu^2}\right), \quad
\mu_{\log} = \log(\mu) - \frac{\sigma_{\log}^2}{2}$$

---

# Roulette Method

The roulette method (Oakley & O'Hagan, 2010) is an interactive chip-allocation
technique that makes elicitation accessible to clinical experts who are
unfamiliar with probability distributions.

The expert places chips in bins representing plausible values. The chip
histogram is then fitted to a parametric family. In the Shiny app, this is
interactive; programmatically:

```{r roulette}
# Expert places chips: 0 very low, rising to peak at 0.25-0.35, falling off.
# Note: chips must not sum such that any cumulative frequency equals exactly
# 0 or 1 — elicit_roulette() strips those boundary bins automatically, but
# at least 2 interior quantile points must remain after stripping.
prior_rou <- elicit_roulette(
  chips  = c(0L, 1L, 3L, 7L, 9L, 7L, 4L, 2L, 1L, 1L),
  breaks = seq(0, 1, by = 0.1),
  family = "beta",
  label  = "Response rate"
)
print(prior_rou)
plot(prior_rou)
```

The roulette object also stores the raw chip data for provenance:

```{r roulette-chips}
prior_rou$roulette$chips
prior_rou$roulette$breaks
```

---

# Expert Pooling

When multiple experts are elicited, their priors are combined into a single
consensus distribution.

## Linear Pooling (Recommended)

Linear pooling produces a weighted mixture — the standard approach in clinical
trials (externally Bayesian):

```{r linear-pooling}
e1 <- elicit_beta(mean = 0.25, sd = 0.08, method = "moments",
                  expert_id = "Oncologist_1", label = "ORR")
e2 <- elicit_beta(mean = 0.35, sd = 0.10, method = "moments",
                  expert_id = "Oncologist_2", label = "ORR")
e3 <- elicit_beta(
  quantiles = c("0.10" = 0.15, "0.50" = 0.30, "0.90" = 0.52),
  expert_id = "Statistician", label = "ORR"
)

consensus <- aggregate_experts(
  priors  = list(
    Oncologist_1 = e1,
    Oncologist_2 = e2,
    Statistician = e3
  ),
  weights = c(0.40, 0.35, 0.25),
  method  = "linear"
)
print(consensus)
plot(consensus)
```

The Bhattacharyya coefficient matrix quantifies pairwise agreement:

```{r bc-matrix}
round(consensus$aggregation$disagreement, 3)
```

Values close to 1 indicate high agreement; values below 0.5 flag substantial
disagreement and should trigger re-elicitation discussion.

## Logarithmic Pooling

Produces a sharper consensus when experts agree, but can be severely influenced
by outlying opinions:

```{r log-pooling}
log_consensus <- aggregate_experts(
  priors  = list(E1 = e1, E2 = e2, E3 = e3),
  weights = c(1/3, 1/3, 1/3),
  method  = "logarithmic"
)
```

## Manual Mixture Construction

For full control over component distributions:

```{r mixture}
mix <- elicit_mixture(
  components = list(e1, e2),
  weights    = c(0.6, 0.4),
  label      = "Weighted two-expert consensus"
)
plot(mix)
```

---

# Prior Object Structure

Every elicited prior is a `bayprior` S3 object with a consistent structure:

```{r structure}
str(e1, max.level = 2)
```

Key fields:

| Field | Content |
|---|---|
| `$dist` | Distribution family (`"beta"`, `"normal"`, etc.) |
| `$params` | Named list of hyperparameters |
| `$method` | Elicitation method used |
| `$expert_id` | Expert identifier for provenance |
| `$label` | Human-readable quantity description |
| `$fit_summary` | Mean, SD, and 2.5/50/97.5th percentiles |
| `$input` | Raw elicitation inputs (quantiles or moments) |

---

# References

O'Hagan, A., Buck, C. E., Daneshkhah, A., Eiser, J. R., Garthwaite, P. H.,
Jenkinson, D. J., Oakley, J. E., & Rakow, T. (2006). *Uncertain Judgements:
Eliciting Experts' Probabilities*. Wiley.

Oakley, J. E. & O'Hagan, A. (2010). SHELF: the Sheffield Elicitation
Framework. University of Sheffield.
