---
title: "Regulatory Reporting"
author: "bayprior package"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Regulatory Reporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 6,
  fig.height = 3.8,
  out.width  = "95%",
  dpi        = 120,
  warning    = FALSE,
  message    = FALSE
)
library(bayprior)
```

# Regulatory Context

The FDA Draft Guidance on Bayesian Statistical Methods for Drug and Biological
Products (2026) and the EMA Reflection Paper on Bayesian Statistics both
require that Bayesian prior distributions used in confirmatory clinical trials
be:

1. **Explicitly specified** — family, hyperparameters, and elicitation method
2. **Justified** — documented basis (expert opinion, historical data, or both)
3. **Conflict-checked** — evidence that prior and data are compatible
4. **Sensitivity-tested** — conclusions hold under plausible alternative priors
5. **Pre-specified** — agreed with regulators before unblinding

`prior_report()` generates a self-contained document addressing all five
requirements.

---

# The Prior Justification Report

## What It Contains

A bayprior report includes:

- **Executive summary** — one-page overview for non-statistical reviewers
- **Trial information** — protocol number, sponsor, statistician, date
- **Prior specification** — distribution, hyperparameters, summary statistics,
  density plot, and elicitation method
- **Conflict diagnostics** — Box p-value, surprise index, KL divergence,
  Bhattacharyya overlap, severity classification, and recommendation
- **Prior-likelihood-posterior overlay** — the canonical visual for conflict
- **Sensitivity analysis** — tornado plot and influence heatmap
- **Regulatory compliance checklist** — maps to FDA/EMA requirements
- **Session information** — R version, bayprior version, Quarto CLI version

## Output Formats

Three formats are supported:

| Format | Use case |
|---|---|
| HTML | Internal review, version-controlled documentation |
| PDF | Regulatory submission appendix |
| Word (.docx) | Co-authoring and track-changes review |

---

# Generating a Report

```{r eval = FALSE}
# 1. Elicit the prior
prior <- elicit_beta(
  mean      = 0.30,
  sd        = 0.10,
  method    = "moments",
  expert_id = "Expert_1",
  label     = "Objective response rate"
)

# 2. Run conflict diagnostics
cd <- prior_conflict(
  prior        = prior,
  data_summary = list(type = "binary", x = 18, n = 40),
  alpha        = 0.05
)

# 3. Run sensitivity analysis
sa <- sensitivity_grid(
  prior        = prior,
  data_summary = list(type = "binary", x = 14, n = 40),
  param_grid   = list(alpha = seq(1, 8, 0.5), beta = seq(2, 20, 1)),
  target       = c("posterior_mean", "prob_efficacy"),
  threshold    = 0.30
)

# 4. Generate the report
prior_report(
  prior         = prior,
  conflict      = cd,
  sensitivity   = sa,
  output_format = "pdf",
  trial_name    = "TRIAL-001",
  sponsor       = "BioPharma Ltd",
  author        = "J. Smith, Principal Biostatistician",
  notes         = paste0(
    "Prior elicited from two independent oncologists (Expert_1, Expert_2) ",
    "using the SHELF structured elicitation protocol. Experts were blinded ",
    "to interim data at the time of elicitation. Prior was pre-specified in ",
    "the Statistical Analysis Plan dated 2025-09-01."
  )
)
```

---

# The Notes Field

The `notes` argument is the statistician's narrative — the most important
section for regulatory reviewers because it provides the scientific rationale
that no automated output can supply. It should address:

**1. Basis for the prior**

```
"Prior elicited from three independent clinical experts in non-small cell
lung cancer using SHELF quantile matching (O'Hagan et al., 2006). Experts
were asked to specify the 10th, 50th, and 90th percentiles of the ORR
distribution based on knowledge of similar agents in this indication."
```

**2. Expert independence and blinding**

```
"All experts were blinded to any interim results at the time of elicitation.
No expert had a financial interest in the trial outcome."
```

**3. Pre-specification**

```
"The prior distribution and elicitation protocol were pre-specified in the
Bayesian Statistical Analysis Plan (BSAP), version 2.1, submitted to FDA
on 2025-09-01 as part of the IND amendment."
```

**4. Conflict action plan**

```
"In the event of mild conflict (Box p-value 0.01-0.05), a sensitivity
analysis using the robust mixture prior (vague weight 0.30) will be added.
In the event of severe conflict (Box p-value < 0.01), the primary analysis
will be repeated using the sceptical prior as a co-primary analysis."
```

---

# The FDA/EMA Compliance Checklist

The report automatically generates a compliance checklist that maps each
section to specific regulatory requirements:

```{r checklist-demo}
library(knitr)
kable(data.frame(
  Requirement = c(
    "Prior elicitation method documented",
    "Distribution family and parameters specified",
    "Expert / source of prior identified",
    "Prior density plot provided",
    "Prior-data conflict assessed",
    "Conflict diagnostic statistics reported",
    "Sensitivity analysis performed",
    "Sensitivity visualisations provided",
    "Alternative priors considered",
    "Regulatory report generated"
  ),
  `FDA Guidance Section` = c(
    "Section IV.B", "Section IV.B", "Section IV.B",
    "Section IV.B", "Section IV.C", "Section IV.C",
    "Section IV.D", "Section IV.D", "Section IV.D",
    "Section IV.E"
  ),
  check.names = FALSE
), align = "ll")
```

---

# Session Reproducibility

The Session Information section of every report records the full computational
environment:

| Item | Purpose |
|---|---|
| R version | Core language version |
| bayprior version | Package version — pin this in your SAP |
| quarto R package | Report rendering R interface |
| Quarto CLI version | Actual rendering engine version |
| Platform | Operating system and architecture |
| Date | Date of report generation |

For fully reproducible regulatory submissions, pin the bayprior version in
your `renv.lock` file and archive the complete `renv` snapshot alongside the
submission documents.

---

# Working with the Shiny App

The `run_app()` function launches the full interactive Shiny interface. Every
analysis conducted in the app — elicitation, pooling, conflict diagnostics,
and sensitivity — is preserved in the session state and flows directly into
the report when the Download button is clicked.

Figures in the report reflect exactly what the analyst saw in the app at the
time of download. The app captures ggplot objects from each panel at the
moment the Download button is pressed, so the report is a faithful snapshot
of the specific analysis session.

```{r eval = FALSE}
# Launch the interactive app
run_app()
```

The app workflow mirrors the six-step analytical workflow:

```
Welcome → Prior Elicitation → Expert Pooling → Conflict Diagnostics →
Sensitivity Analysis → Robust Priors → Export Report
```

---

# Pre-Submission Checklist

Before submitting a bayprior report to a regulatory agency:

- [ ] Prior was pre-specified in the Statistical Analysis Plan
- [ ] Expert elicitation was conducted blind to interim data
- [ ] Expert identities and conflicts of interest are documented
- [ ] Conflict action plan is pre-specified (what happens at each severity)
- [ ] Sensitivity range was agreed with regulators a priori
- [ ] At least one alternative prior is reported (robust or sceptical)
- [ ] bayprior version is pinned in the renv.lock
- [ ] Report generated from installed package (not `load_all()`)
- [ ] PDF and HTML versions both generated and cross-checked
- [ ] Notes field contains full scientific rationale

---

# References

FDA (2026). *Draft Guidance for Industry: Bayesian Statistical Methods for
Drug and Biological Products*.

EMA (2023). *Reflection Paper on the Use of Bayesian Statistics in the
Evaluation of Medicinal Products*.

O'Hagan, A. et al. (2006). *Uncertain Judgements: Eliciting Experts'
Probabilities*. Wiley.

Spiegelhalter, D. J. & Freedman, L. S. (1994). Bayesian approaches to
clinical trials. *JRSS-A*, 157, 357–416.
