---
title: "Robust, Sceptical, and Power Priors"
author: "bayprior package"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Robust, Sceptical, and Power Priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 6,
  fig.height = 3.8,
  out.width  = "95%",
  dpi        = 120,
  warning    = FALSE,
  message    = FALSE
)
library(bayprior)
```

# Overview

Beyond the elicited informative prior, regulatory submissions typically require
one or more alternative prior specifications to demonstrate robustness. bayprior
provides three well-established alternatives:

| Prior type | Function | Method | Reference |
|---|---|---|---|
| **Robust mixture** | `robust_prior()` | Mixes informative + vague | Schmidli et al. (2014) |
| **Sceptical** | `sceptical_prior()` | Centred at null effect | Spiegelhalter et al. (1994) |
| **Power prior** | `calibrate_power_prior()` | Down-weights historical data | Ibrahim & Chen (2000) |

---

# Robust Mixture Prior

## Concept

A robust prior protects against prior misspecification by mixing the informative
elicited prior with a vague (diffuse) component:

$$\pi_{\text{robust}}(\theta) = (1 - w) \cdot \pi_{\text{informative}}(\theta)
+ w \cdot \pi_{\text{vague}}(\theta)$$

The vague component — a wide Normal centred at the informative prior mean —
ensures the posterior is never completely dominated by a conflicting prior.
The default weight is $w = 0.20$ (80% informative, 20% vague).

## Usage

```{r robust}
informative <- elicit_beta(
  mean      = 0.30,
  sd        = 0.08,
  method    = "moments",
  label     = "Response rate"
)

rob <- robust_prior(
  informative  = informative,
  vague_weight = 0.20,
  label        = "Robust mixture prior"
)
plot(rob)
```

```{r robust-summary}
cat("Informative component weight:", 1 - rob$vague_weight, "\n")
cat("Vague component weight:      ", rob$vague_weight, "\n")
cat("Mixture mean:", round(rob$fit_summary$mean, 4), "\n")
cat("Mixture SD:  ", round(rob$fit_summary$sd,   4), "\n")
```

## Varying the Vague Weight

A higher vague weight makes the prior more diffuse and reduces its influence
on the posterior:

```{r robust-weights, fig.height = 4}
par(mfrow = c(1, 1))
weights <- c(0.10, 0.20, 0.30, 0.50)
cols    <- c("#185FA5", "#1D9E75", "#D85A30", "#888780")

x <- seq(0, 0.8, length.out = 300)
plot(x, bayprior:::.eval_density_vec(informative, x),
     type = "l", lwd = 2, col = "#185FA5",
     xlab = "Response rate", ylab = "Density",
     main = "Effect of vague weight on robust prior",
     ylim = c(0, 6))
for (i in seq_along(weights)) {
  r <- robust_prior(informative, vague_weight = weights[i])
  lines(x, bayprior:::.eval_density_vec(r, x),
        col = cols[i], lwd = 1.5, lty = i + 1)
}
legend("topright",
       legend = c("Informative",
                  paste0("w = ", weights)),
       col    = c("#185FA5", cols),
       lwd    = 2,
       lty    = c(1, 2, 3, 4, 5),
       bty    = "n", cex = 0.8)
```

## Vague SD Multiplier

The vague component's SD defaults to 10× the informative prior's SD. Adjust
with `vague_sd`:

```{r robust-sd}
rob_narrow <- robust_prior(informative, vague_weight = 0.20,
                           vague_sd = 2 * informative$fit_summary$sd)
rob_wide   <- robust_prior(informative, vague_weight = 0.20,
                           vague_sd = 20 * informative$fit_summary$sd)

cat("Narrow vague SD: ", round(rob_narrow$components$vague$fit_summary$sd, 3), "\n")
cat("Default vague SD:", round(rob$components$vague$fit_summary$sd, 3), "\n")
cat("Wide vague SD:   ", round(rob_wide$components$vague$fit_summary$sd, 3), "\n")
```

---

# Sceptical Prior

## Concept

The sceptical prior (Spiegelhalter & Freedman, 1994) represents the view of a
conservative regulator who is sceptical of a treatment effect. It is centred at
the null value of the treatment effect with width calibrated to a chosen
strength of scepticism.

This is the FDA's recommended sensitivity prior for trials using informative
priors: the trial conclusions should hold even under a prior that places most
mass at "no effect."

## Normal Family (Mean Differences, Log Odds Ratios)

For continuous or log-scale quantities where the null is typically 0:

```{r sceptical-normal}
sc_weak <- sceptical_prior(
  null_value = 0, family = "normal", strength = "weak",
  label = "Log OR (weak sceptic)"
)
sc_moderate <- sceptical_prior(
  null_value = 0, family = "normal", strength = "moderate",
  label = "Log OR (moderate sceptic)"
)
sc_strong <- sceptical_prior(
  null_value = 0, family = "normal", strength = "strong",
  label = "Log OR (strong sceptic)"
)

cat("Weak SD:    ", sc_weak$fit_summary$sd, "\n")
cat("Moderate SD:", sc_moderate$fit_summary$sd, "\n")
cat("Strong SD:  ", sc_strong$fit_summary$sd, "\n")
```

```{r sceptical-normal-plot}
plot(sc_moderate)
```

The SD mapping by strength:

| Strength | SD | Interpretation |
|---|---|---|
| `weak` | 1.0 | Vague scepticism — wide prior around null |
| `moderate` | 0.5 | 2-SD departure from null has ~5% prior probability |
| `strong` | 0.25 | Very concentrated at null — very sceptical |

## Beta Family (Response Rates)

For binary endpoints, `null_value` must be in $(0, 1)$ — it represents the null
response rate, not a difference:

```{r sceptical-beta}
# Null response rate of 20%: sceptic believes treatment is no better than 20%
sc_beta <- sceptical_prior(
  null_value = 0.20,
  family     = "beta",
  strength   = "moderate",
  label      = "Response rate (sceptical)"
)
plot(sc_beta)
```

## Log-Normal Family (Hazard Ratios)

For hazard ratios, the null is HR = 1, which corresponds to `null_value = 0`
on the log scale:

```{r sceptical-lognormal}
sc_hr <- sceptical_prior(
  null_value = 0,         # log(1) = 0, i.e. HR = 1
  family     = "lognormal",
  strength   = "moderate",
  label      = "Hazard ratio (sceptical)"
)
plot(sc_hr)
```

## Enthusiastic vs Sceptical Pair

The FDA recommends presenting conclusions under both an enthusiastic prior
(favouring treatment benefit) and a sceptical prior:

```{r enthusiastic-sceptical}
enthusiastic <- elicit_beta(
  mean = 0.45, sd = 0.08,
  method = "moments", label = "Response rate (enthusiastic)"
)
sceptical <- sceptical_prior(
  null_value = 0.20, family = "beta", strength = "moderate",
  label = "Response rate (sceptical)"
)

data_obs <- list(type = "binary", x = 18, n = 40)

post_enth <- bayprior:::.conjugate_update(enthusiastic, data_obs)
post_scep <- bayprior:::.conjugate_update(sceptical,    data_obs)

cat("Posterior mean (enthusiastic):", round(post_enth$fit_summary$mean, 3), "\n")
cat("Posterior mean (sceptical):   ", round(post_scep$fit_summary$mean, 3), "\n")
cat("Posterior SD (enthusiastic):  ", round(post_enth$fit_summary$sd,   3), "\n")
cat("Posterior SD (sceptical):     ", round(post_scep$fit_summary$sd,   3), "\n")
```

---

# Power Prior

## Concept

The power prior (Ibrahim & Chen, 2000) provides a principled method for
incorporating historical data by down-weighting it by a factor
$\delta \in (0, 1]$:

$$\pi(\theta | D_0, \delta) \propto L(\theta | D_0)^\delta \cdot \pi_0(\theta)$$

where $D_0$ is the historical data and $\delta$ controls how much weight it
receives. $\delta = 1$ fully incorporates the historical data (standard
Bayesian updating); $\delta \to 0$ ignores it entirely.

## Calibrating $\delta$

`calibrate_power_prior()` selects $\delta$ to achieve a target Bayes Factor
between the historical-data-informed prior and the current likelihood —
ensuring the historical data is incorporated only to the extent it is
compatible with current data:

```{r power-prior, cache = TRUE}
base <- elicit_beta(
  mean   = 0.50,
  sd     = 0.20,
  method = "moments",
  label  = "Response rate"
)

calib <- calibrate_power_prior(
  historical_data = list(type = "binary", x = 12, n = 40),
  current_data    = list(type = "binary", x = 18, n = 50),
  base_prior      = base,
  target_bf       = 3,
  delta_grid      = seq(0.05, 1.0, by = 0.05),
  method          = "bayes_factor"
)
print(calib)
```

```{r power-plot}
plot(calib)
```

The calibration curves show:

- **Top panel:** Bayes Factor vs $\delta$. The dashed red line is the target BF;
  the dotted green line marks the optimal $\delta$.
- **Bottom panel:** Box p-value vs $\delta$. Values below 0.05 indicate
  conflict at that weight.

## Compatibility Method

Alternatively, select $\delta$ to be the largest value for which the
historical-data-informed prior shows no conflict with the current data:

```{r power-compat, cache = TRUE}
calib_compat <- calibrate_power_prior(
  historical_data = list(type = "binary", x = 12, n = 40),
  current_data    = list(type = "binary", x = 18, n = 50),
  base_prior      = base,
  method          = "compatibility",
  delta_grid      = seq(0.05, 1.0, by = 0.05)
)
cat("Optimal delta (BF method):           ", calib$delta_opt, "\n")
cat("Optimal delta (compatibility method):", calib_compat$delta_opt, "\n")
```

## Power Prior for Other Families

Power prior updating is supported for Beta, Normal, Gamma, Log-Normal, and
Mixture priors. For a Normal prior with continuous historical data:

```{r power-normal, cache = TRUE}
base_norm <- elicit_normal(
  mean = 0.0, sd = 0.5,
  method = "moments", label = "Mean difference"
)

calib_norm <- calibrate_power_prior(
  historical_data = list(type = "continuous", x = 0.35, sd = 0.3, n = 60),
  current_data    = list(type = "continuous", x = 0.42, sd = 0.3, n = 80),
  base_prior      = base_norm,
  target_bf       = 3,
  delta_grid      = seq(0.05, 1.0, by = 0.10),
  method          = "bayes_factor"
)
print(calib_norm)
```

---

# Choosing the Right Alternative Prior

```{r choice-table}
library(knitr)
kable(data.frame(
  Situation = c(
    "No conflict, regulatory requirement",
    "Mild conflict detected",
    "Severe conflict detected",
    "Historical data available",
    "FDA enthusiastic/sceptical pair required"
  ),
  `Recommended prior` = c(
    "Robust mixture (w = 0.20)",
    "Robust mixture (w = 0.30-0.40)",
    "Sceptical prior (moderate-strong)",
    "Power prior (calibrated)",
    "Sceptical prior as second arm"
  ),
  check.names = FALSE
), align = "ll")
```

---

# References

Ibrahim, J. G. & Chen, M.-H. (2000). Power prior distributions for regression
models. *Statistical Science*, 15, 46–60.

Schmidli, H., Gsteiger, S., Roychoudhury, S., O'Hagan, A., Spiegelhalter, D.,
& Neuenschwander, B. (2014). Robust meta-analytic-predictive priors in clinical
trials with historical control information. *Biometrics*, 70, 1023–1032.

Spiegelhalter, D. J., Freedman, L. S., & Parmar, M. K. B. (1994). Bayesian
approaches to randomized trials. *Journal of the Royal Statistical Society A*,
157, 357–416.

Gravestock, I. & Held, L. (2017). Adaptive power priors with empirical Bayes
for clinical trials. *Pharmaceutical Statistics*, 16, 349–360.
