---
title: "Sensitivity Analysis"
author: "bayprior package"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Sensitivity Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 6,
  fig.height = 3.8,
  out.width  = "95%",
  dpi        = 120,
  warning    = FALSE,
  message    = FALSE
)
library(bayprior)
```

# Why Sensitivity Analysis Is Required

Regulatory guidelines — including the FDA Draft Guidance on Bayesian
Statistical Methods (2026) and ICH E9(R1) — require demonstrating that trial
conclusions are robust to plausible variations in prior specification. A
sensitivity analysis answers the question: *How much do posterior conclusions
change if the prior hyperparameters are varied across their plausible range?*

**bayprior** provides two dedicated sensitivity functions:

- `sensitivity_grid()` — evaluates posterior mean, SD, and efficacy probability
- `sensitivity_cri()` — focuses specifically on credible interval width

---

# Setup

```{r setup-prior}
prior <- elicit_beta(
  mean   = 0.30,
  sd     = 0.10,
  method = "moments",
  label  = "Response rate"
)

data_obs <- list(type = "binary", x = 14, n = 40)
```

---

# `sensitivity_grid()`

## Basic Usage

Evaluates how posterior quantities change as hyperparameters vary over a
Cartesian grid. The `param_grid` names must match the prior's hyperparameter
names (`alpha` and `beta` for a Beta prior):

```{r sensitivity-basic, cache = TRUE}
sa <- sensitivity_grid(
  prior        = prior,
  data_summary = data_obs,
  param_grid   = list(alpha = seq(1, 8, 1), beta = seq(2, 20, 2)),
  target       = c("posterior_mean", "prob_efficacy"),
  threshold    = 0.30
)
```

The result is a `bayprior_sensitivity` object:

```{r sensitivity-str}
names(sa)
sa$influence_scores
```

**Influence scores** are the range of each target quantity across the full
hyperparameter grid — the primary regulatory summary statistic.

## Tornado Plot

Ranks posterior quantities from most to least sensitive. Bar width equals the
influence score. The blue dot marks the reference prior (closest grid point to
the elicited hyperparameters):

```{r tornado, fig.height = 3}
plot_tornado(sa)
```

## Influence Heatmap

Shows how the posterior mean changes across the two-dimensional hyperparameter
grid. Useful for identifying which region of the prior space drives sensitivity:

```{r heatmap-mean, fig.height = 4}
plot_sensitivity(sa, target = "posterior_mean")
```

```{r heatmap-efficacy, fig.height = 4}
plot_sensitivity(sa, target = "prob_efficacy")
```

The orange diamond marks the reference prior. A well-specified prior should sit
in a region of low colour gradient — indicating that small hyperparameter
perturbations produce little change in posterior conclusions.

---

# `sensitivity_cri()`

Tracks the width and bounds of the posterior credible interval (CrI) — a key
quantity for regulatory submissions because it directly addresses whether the
CrI for the treatment effect remains favourable across plausible prior
specifications:

```{r cri-sensitivity, cache = TRUE}
cri_sa <- sensitivity_cri(
  prior        = prior,
  data_summary = data_obs,
  param_grid   = list(alpha = seq(1, 8, 1), beta = seq(2, 20, 2)),
  cri_level    = 0.95,
  threshold    = 0.30
)
```

```{r cri-width, fig.height = 4}
plot_sensitivity(cri_sa, target = "cri_width")
```

```{r cri-lower, fig.height = 4}
plot_sensitivity(cri_sa, target = "cri_lower")
```

The `cri_sa$influence_scores` gives the range of each CrI quantity:

```{r cri-scores}
round(cri_sa$influence_scores, 4)
```

---

# Interpreting Influence Scores

```{r interpretation-table}
library(knitr)
kable(data.frame(
  Score     = c("< 0.02", "0.02 - 0.05", "0.05 - 0.10", "> 0.10"),
  Posterior = c("Negligible sensitivity",
                "Low sensitivity — prior well-specified",
                "Moderate sensitivity — consider reporting",
                "High sensitivity — investigate further"),
  check.names = FALSE
), align = "ll")
```

A posterior mean influence score of 0.03 means the posterior mean shifts by at
most 3 percentage points across the entire plausible hyperparameter range — a
regulatory-defensible result for most oncology trials.

---

# Sensitivity for Other Prior Families

## Normal Prior

For a continuous endpoint with a Normal prior:

```{r normal-sensitivity, cache = TRUE}
prior_norm <- elicit_normal(
  mean   = 0.0,
  sd     = 0.3,
  method = "moments",
  label  = "Log odds ratio"
)

sa_norm <- sensitivity_grid(
  prior        = prior_norm,
  data_summary = list(type = "continuous", x = 0.4, sd = 0.3, n = 60),
  param_grid   = list(mu = seq(-0.5, 0.5, 0.25), sigma = seq(0.1, 0.6, 0.1)),
  target       = c("posterior_mean", "posterior_sd")
)
plot_tornado(sa_norm)
```

## Mixture Prior

For mixture priors, `sensitivity_grid()` internally uses the dominant component
for the hyperparameter grid — avoiding the complexity of gridding over mixture
weights:

```{r mixture-sensitivity, cache = TRUE}
e1 <- elicit_beta(mean = 0.25, sd = 0.08, method = "moments",
                  expert_id = "E1", label = "Response rate")
e2 <- elicit_beta(mean = 0.40, sd = 0.10, method = "moments",
                  expert_id = "E2", label = "Response rate")
mix <- aggregate_experts(
  priors  = list(E1 = e1, E2 = e2),
  weights = c(0.6, 0.4)
)

sa_mix <- sensitivity_grid(
  prior        = mix,
  data_summary = data_obs,
  param_grid   = list(alpha = seq(1, 8, 1), beta = seq(2, 20, 2)),
  target       = c("posterior_mean", "prob_efficacy"),
  threshold    = 0.30
)
plot_tornado(sa_mix)
```

---

# The Reference Row

The reference row is the grid point closest (in Euclidean distance) to the
elicited prior's hyperparameters. It serves as the anchor for interpreting the
heatmap and influence scores:

```{r reference}
# Which grid row is the reference?
sa$reference_row

# What are its parameter values?
sa$grid[sa$reference_row, c("alpha", "beta")]
```

---

# Regulatory Reporting of Sensitivity

The standard regulatory narrative structure is:

1. **State the reference prior** — hyperparameters and elicitation method
2. **Define the sensitivity range** — how far hyperparameters were varied and why
3. **Report influence scores** — for each target quantity
4. **State the conclusion** — whether trial conclusions hold across the range

Example text:

> "Sensitivity of the posterior response rate estimate to prior
> hyperparameter specification was assessed by varying $\alpha$ from 1 to 8
> and $\beta$ from 2 to 20. The posterior mean influence score was
> `r round(sa$influence_scores["posterior_mean"], 3)`, indicating that the
> posterior mean shifts by at most
> `r round(sa$influence_scores["posterior_mean"] * 100, 1)`
> percentage points across the full sensitivity range. Trial conclusions
> are robust to the choice of prior hyperparameters within the pre-specified
> plausible range."

---

# References

FDA (2026). *Draft Guidance: Bayesian Statistical Methods for Drug and
Biological Products*.

ICH E9(R1) (2019). *Statistical Principles for Clinical Trials: Addendum on
Estimands and Sensitivity Analysis*.
