---
title: "bayprior: Structured Bayesian Prior Elicitation, Conflict Diagnostics, and Regulatory Reporting"
date: "`r Sys.Date()`"
abstract: >
  Bayesian clinical trials require a well-specified prior distribution, yet
  no unified R tool previously addressed the full workflow of prior
  construction, validation, and regulatory justification. We introduce
  bayprior, an R package and Shiny application that implements structured
  expert elicitation via quantile matching, moment matching, and the SHELF
  roulette method across six distribution families; linear and logarithmic
  expert opinion pooling with Bhattacharyya agreement diagnostics;
  prior-data conflict assessment using the Box p-value, surprise index,
  information divergence, Bhattacharyya overlap, and multivariate
  Mahalanobis distance; hyperparameter sensitivity analysis with tornado
  and influence heatmap visualisations; and robust, sceptical, and power
  prior alternatives. A regulatory reporting module generates
  self-contained HTML, PDF, or Word documents aligned with FDA and EMA
  guidance on Bayesian methods. The package is demonstrated on a synthetic
  oncology Phase II trial. bayprior is available on CRAN and includes a
  fully modular Shiny application for interactive use.
draft: true
author:
  - name: Ndoh Penn
    affiliation: Independent Researcher
    address:
    - Brussels, Belgium
    url: https://github.com/ndohpenngit
    orcid: 0009-0003-9054-465X
    email: ndohpenn9@gmail.com
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  out.width  = "100%",
  comment    = "#>"
)
library(bayprior)
```

# Introduction

Bayesian statistical methods are increasingly used in clinical trial design
and analysis, offering flexible incorporation of prior knowledge and natural
uncertainty quantification [@FDA2026]. However, the widespread adoption of
Bayesian methods in regulatory settings requires that prior distributions be
not only well-chosen but also transparently documented and demonstrably robust.
The FDA's 2026 draft guidance on Bayesian methods [@FDA2026] explicitly
requires sponsors to demonstrate that trial conclusions hold under
sensitivity analyses over the prior specification, yet provides no
methodological framework for how this should be done.

In practice, prior specification often proceeds informally. Analysts may
elicit a prior from clinical literature or expert opinion without a
structured procedure, assess prior-data conflict informally by visual
inspection, and document the justification in narrative form. This approach
is difficult to reproduce and rarely satisfies regulatory reviewers who
expect a systematic, auditable justification.

Existing R packages for Bayesian clinical trials focus on posterior
computation and adaptive design. \CRANpkg{RBesT} [@RBesT] implements
robust meta-analytic-predictive priors for historical data borrowing.
\CRANpkg{trialr} [@trialr] provides a range of adaptive trial designs
via Stan. \CRANpkg{hdbayes} [@hdbayes] implements multiple historical
data borrowing priors -- including power priors, normalised power priors,
robust MAP priors, and commensurate priors -- for generalised linear
models. None of these packages addresses the upstream problem of prior
construction from expert knowledge, prior-data conflict assessment, or
regulatory reporting.

We introduce \CRANpkg{bayprior}, an R package and Shiny application that
fills this gap by providing a complete prior justification workflow:
structured expert elicitation, expert opinion pooling, prior-data conflict
diagnostics, sensitivity analysis, robust and regulatory prior alternatives,
and automated regulatory reporting. The package implements established
statistical methodology [@OHagan2006; @Box1980; @Oakley2010; @Schmidli2014;
@Ibrahim2000; @Spiegelhalter1994] in a unified, accessible framework
suitable for both programmatic use and interactive exploration via the
bundled Shiny application.

# Related Software

## Expert Elicitation

The Sheffield Elicitation Framework [@Oakley2010] provides a structured
approach to expert elicitation and is implemented in the R package
\CRANpkg{SHELF} [@SHELF]. \CRANpkg{SHELF} supports elicitation from
individual or multiple experts across a range of distribution families,
with weighted linear pooling for aggregating multiple expert beliefs.
\CRANpkg{bayprior} extends this with three programmatic elicitation
methods (quantile matching, moment matching, and the roulette method)
across six distribution families, adds logarithmic opinion pooling as
a complement to linear pooling, and introduces quantitative agreement
diagnostics via the Bhattacharyya coefficient for assessing
inter-expert consistency before pooling.

## Robust Priors and Historical Borrowing

\CRANpkg{RBesT} [@RBesT] implements the robust meta-analytic-predictive
(MAP) prior approach [@Schmidli2014] for incorporating historical control
data. \CRANpkg{hdbayes} [@hdbayes] provides a broader range of historical
borrowing methods, including power priors, normalised power priors,
commensurate priors, and robust MAP priors, for generalised linear models
via Stan. \CRANpkg{bayprior} provides the Schmidli robust mixture prior
and calibrated power priors [@Ibrahim2000] as alternatives within the prior
justification workflow, but focuses on prior specification and justification
rather than historical data meta-analysis.

## Prior Sensitivity and Conflict Diagnostics

\CRANpkg{priorsense} [@priorsense] implements power-scaling sensitivity
analysis for detecting prior-data conflict and likelihood noninformativity
in fitted Stan models [@Kallioinen2023]. \CRANpkg{bayprior} addresses a
complementary use case: closed-form conflict diagnostics for conjugate
prior families at the prior specification stage, before any model is
fitted, without requiring Stan. The two packages serve different points
in the Bayesian workflow.

## Scope Comparison

Table \@ref(tab:pkg-comparison) summarises the scope of existing packages
relative to \CRANpkg{bayprior}.

```{r pkg-comparison, echo=FALSE}
knitr::kable(
  data.frame(
    Package     = c("SHELF", "RBesT", "trialr", "hdbayes",
                    "priorsense", "bayprior"),
    Elicitation = c("Yes",     "No",    "No",     "No",
                    "No",       "Yes"),
    Pooling     = c("Partial", "No",    "No",     "No",
                    "No",       "Yes"),
    Conflict    = c("No",      "No",    "No",     "No",
                    "Yes",      "Yes"),
    Sensitivity = c("No",      "No",    "No",     "No",
                    "Yes",      "Yes"),
    `Robust priors` = c("No",  "Yes",   "No",     "Yes",
                        "No",   "Yes"),
    Reporting   = c("No",      "No",    "No",     "No",
                    "No",       "Yes"),
    check.names = FALSE
  ),
  caption = "Scope comparison of R packages for Bayesian prior specification.
             Partial: SHELF pooling supports linear pooling only.
             priorsense conflict and sensitivity diagnostics require a
             fitted Stan model.",
  booktabs = TRUE
)
```

# Package Overview

\CRANpkg{bayprior} is built using the \CRANpkg{golem} [@golem] framework
and provides both exported R functions for programmatic use and a modular
Shiny application for interactive exploration. The package is available on
CRAN and can be installed with:

```{r install, eval=FALSE}
install.packages("bayprior")

# Or the development version from GitHub:
remotes::install_github("ndohpenngit/bayprior")
```

The Shiny application can be launched with:

```{r run-app, eval=FALSE}
library(bayprior)
run_app()
```

## Prior Elicitation

\CRANpkg{bayprior} implements three structured elicitation approaches
across six distribution families (Beta, Normal, Gamma, Log-Normal,
Exponential, and Weibull).

**Moment matching** derives hyperparameters analytically from an
expert-supplied mean and standard deviation:

```{r elicitation-moments}
prior <- elicit_beta(
  mean      = 0.35,
  sd        = 0.10,
  method    = "moments",
  label     = "Response rate",
  expert_id = "Expert_1"
)
print(prior)
```

**Quantile matching** fits the distribution numerically to expert-specified
probability--value pairs, accommodating asymmetric beliefs that moment
matching cannot capture. For example, an expert who believes the response
rate has a 25\% chance of being below 0.20, a 60\% chance of being below
0.35, and an 80\% chance of being below 0.50:

```{r elicitation-quantile}
prior_qm <- elicit_beta(
  probs     = c(0.25, 0.60, 0.80),
  quantiles = c(0.20, 0.35, 0.50),
  method    = "quantile",
  label     = "Response rate (quantile)",
  expert_id = "Expert_1"
)
```

```{r elicitation-plot, fig.cap="Beta prior elicited by moment matching (mean = 0.35, SD = 0.10). Shaded region shows the 95\\% credible interval."}
plot(prior)
```

The fitted distribution, parameter estimates, and a 95\% credible interval
are displayed automatically. All six families share a common API, so
`elicit_beta()` can be replaced with `elicit_normal()`, `elicit_gamma()`,
`elicit_lognormal()`, `elicit_exponential()`, or `elicit_weibull()` with
identical syntax.

## Expert Pooling

Multiple expert priors can be aggregated using linear or logarithmic
opinion pooling [@OHagan2006]. Linear pooling takes the weighted average
of the densities; logarithmic pooling takes the weighted geometric mean.

```{r pooling}
e1 <- elicit_beta(mean = 0.30, sd = 0.08, method = "moments",
                  label = "Response rate", expert_id = "Expert_1")
e2 <- elicit_beta(mean = 0.42, sd = 0.10, method = "moments",
                  label = "Response rate", expert_id = "Expert_2")

consensus <- aggregate_experts(
  priors  = list(E1 = e1, E2 = e2),
  weights = c(0.5, 0.5),
  method  = "linear"
)
print(consensus)
```

Agreement between experts is quantified by the Bhattacharyya coefficient
[@Bhattacharyya1943], which measures distributional overlap on a scale from
0 (no overlap) to 1 (identical distributions). This provides an objective
basis for assessing whether pooling is appropriate.

## Prior-Data Conflict Diagnostics

Prior-data conflict occurs when the observed data is implausible under the
prior predictive distribution [@Box1980]. \CRANpkg{bayprior} computes four
complementary conflict metrics:

- **Prior predictive p-value** [@Box1980]: the probability of observing
  data at least as extreme as the observed data under the prior predictive
  distribution.
- **Surprise index**: the standardised distance between the prior mean and
  the observed data, expressed in prior standard deviation units.
- **Information divergence**: the Kullback-Leibler divergence from the
  prior to the normalised likelihood.
- **Bhattacharyya overlap**: the distributional overlap between the prior
  and normalised likelihood.

```{r conflict}
interim <- list(type = "binary", x = 18, n = 40)
cd <- prior_conflict(prior, interim)
print(cd)
```

Four data types are supported: binary (Beta-Binomial conjugate),
continuous (Normal-Normal), Poisson/count (Gamma-Poisson), and survival
(Gamma-Exponential).

For co-primary endpoints, multivariate conflict can be assessed using the
Mahalanobis distance [@Mahalanobis1936]:

```{r conflict-mv}
pm   <- c(0.35, 0.60)
pcov <- matrix(c(0.010, 0.003, 0.003, 0.015), 2, 2)
om   <- c(0.52, 0.58)
ocov <- matrix(c(2e-4, 4e-5, 4e-5, 2e-4), 2, 2)

mv <- conflict_mahalanobis(pm, pcov, om, ocov,
                           labels = c("Response rate", "OS rate"))
print(mv)
```

## Sensitivity Analysis

The sensitivity module evaluates posterior conclusions across a
hyperparameter grid, answering the regulatory question: does the
treatment benefit conclusion change materially under plausible alternative
priors?

```{r sensitivity, fig.cap="Tornado plot showing the influence of prior hyperparameters on the posterior mean. Parameters are ranked by their marginal influence."}
sa <- sensitivity_grid(
  prior        = prior,
  data_summary = interim,
  param_grid   = list(
    alpha = seq(1, 8, by = 0.5),
    beta  = seq(2, 20, by = 1)
  ),
  target    = c("posterior_mean", "prob_efficacy"),
  threshold = 0.30
)
plot_tornado(sa)
```

The `sensitivity_cri()` function tracks credible interval width
specifically -- a quantity directly relevant to the precision of treatment
effect estimates in regulatory submissions.

## Robust, Sceptical, and Power Priors

\CRANpkg{bayprior} provides three alternative prior types for regulatory
sensitivity analyses.

**Robust mixture prior** [@Schmidli2014]: mixes the informative prior with
a vague component, protecting against prior misspecification:

```{r robust}
rob <- robust_prior(prior, vague_weight = 0.20)
plot(rob)
```

**Sceptical prior** [@Spiegelhalter1994]: centred at the null value of
the treatment effect, representing a conservative regulatory stance:

```{r sceptical}
scep <- sceptical_prior(
  null_value = 0.20,
  family     = "beta",
  strength   = "moderate",
  label      = "Response rate (sceptical)"
)
```

**Power prior** [@Ibrahim2000]: down-weights historical data by a factor
$\delta \in (0, 1]$ calibrated to achieve a target Bayes Factor:

```{r power-prior}
base  <- elicit_beta(mean = 0.50, sd = 0.20, method = "moments",
                     label = "Response rate")
calib <- calibrate_power_prior(
  historical_data = list(type = "binary", x = 12, n = 40),
  current_data    = list(type = "binary", x = 18, n = 50),
  base_prior      = base,
  target_bf       = 3
)
print(calib)
```

## Regulatory Reporting

A single call to `prior_report()` generates a self-contained prior
justification document in HTML, PDF, or Word format:

```{r report, eval=FALSE}
prior_report(
  prior           = prior,
  conflict        = cd,
  sensitivity     = sa,
  robust_prior    = rob,
  sceptical_prior = scep,
  output_format   = "html",
  trial_name      = "TRIAL-001",
  sponsor         = "Example Pharma Ltd",
  author          = "N.P., Biostatistician"
)
```

The generated report includes parameter summary tables, all diagnostic
plots, a compliance checklist aligned with the FDA 2026 draft guidance
[@FDA2026], and a structured prior justification narrative.

# Case Study: TRIAL-001

We demonstrate the complete \CRANpkg{bayprior} workflow on TRIAL-001, a
synthetic oncology Phase II trial with a binary response rate endpoint.
The target response rate is 30\% (null hypothesis) versus an expected
rate of 40\% (alternative hypothesis).

## Expert Elicitation and Pooling

Two clinical experts independently specified Beta priors for the response
rate. Expert 1 (conservative) placed a mean of 0.30 with SD of 0.08;
Expert 2 (optimistic) placed a mean of 0.42 with SD of 0.10. Equal weights
were assigned on the basis of comparable seniority.

```{r trial-elicitation}
e1 <- elicit_beta(mean = 0.30, sd = 0.08, method = "moments",
                  label = "Response rate", expert_id = "Expert_1")
e2 <- elicit_beta(mean = 0.42, sd = 0.10, method = "moments",
                  label = "Response rate", expert_id = "Expert_2")

pool <- aggregate_experts(list(E1 = e1, E2 = e2),
                          weights = c(0.5, 0.5),
                          method  = "linear")
print(pool)
```

The Bhattacharyya coefficient between the two expert distributions
indicated substantial agreement, supporting the use of equal-weight
pooling. The consensus prior had a mean of
`r round(pool$fit_summary$mean, 3)` and SD of
`r round(pool$fit_summary$sd, 3)`.

## Conflict Diagnostics at Interim

At the pre-specified interim analysis, 18 responses were observed in
40 patients (observed rate: 45\%).

```{r trial-conflict}
interim_data <- list(type = "binary", x = 18, n = 40)
cd_trial <- prior_conflict(pool, interim_data)
print(cd_trial)
```

The Box p-value was above the 0.10 threshold, indicating no evidence of
prior-data conflict. The posterior distribution updated the consensus prior
towards the observed data as expected.

## Regulatory Sensitivity Pair

Following FDA guidance recommendations, we computed an enthusiastic-sceptical
prior sensitivity pair:

```{r trial-sensitivity-pair}
scep_trial <- sceptical_prior(null_value = 0.20, family = "beta",
                              strength = "moderate",
                              label = "Response rate (sceptical)")

cd_scep <- prior_conflict(scep_trial, interim_data)

cat("Conflict severity (consensus): ",
    cd_trial$conflict_severity, "\n")
cat("Conflict severity (sceptical):",
    cd_scep$conflict_severity, "\n")
```

Conclusions were consistent across both priors, supporting regulatory
robustness.

# Discussion

\CRANpkg{bayprior} provides a structured workflow for prior justification
that addresses a practical gap in the Bayesian clinical trial toolkit. The
package is designed to be used alongside existing analysis packages
(\CRANpkg{RBesT}, \CRANpkg{rstanarm}, \CRANpkg{brms}) rather than
replacing them: bayprior handles the prior specification phase, and
established packages handle the posterior computation.

Several limitations apply. The multivariate Mahalanobis check is currently
limited to bivariate (k = 2) endpoints; extension to k $\geq$ 3 is
planned. Survival power priors require numerical integration and are not
yet implemented. The robust mixture prior uses a Normal vague component
regardless of the informative prior family; for Beta informative priors,
the mixture density is computed numerically rather than analytically.

Future development includes survival power priors, k $\geq$ 3 endpoint
Mahalanobis conflict, and interoperability with \CRANpkg{RBesT} for MAP
prior workflows. A companion paper in Pharmaceutical Statistics
(forthcoming) provides a methodological treatment of the conflict
diagnostic framework and a regulatory alignment assessment against the
FDA 2026 draft guidance and ICH E9(R1).

# Summary

\CRANpkg{bayprior} provides an integrated R tool for the complete prior
justification workflow in Bayesian clinical trials, combining structured
expert elicitation, opinion pooling, conjugate conflict diagnostics,
hyperparameter sensitivity analysis, robust prior alternatives, and
regulatory reporting in a single package. The package is available on CRAN:

```{r install-cran, eval=FALSE}
install.packages("bayprior")
```

The full documentation and vignettes are available at
\url{https://ndohpenngit.github.io/bayprior/}. A live Shiny application
is hosted at \url{https://npenn.shinyapps.io/bayprior/}.

```{r citation, echo=FALSE}
cat(format(citation("bayprior"), style = "text"), sep = "\n")
```
