---
title: "bayprior: Structured Bayesian Prior Elicitation, Conflict Diagnostics, and Regulatory Reporting"
date: "2026-08-03"
abstract: >
  Bayesian clinical trials require a well-specified prior distribution, and
  regulators increasingly expect that specification to be documented,
  diagnosed for prior-data conflict, and shown to be robust to reasonable
  alternatives -- yet few R tools connect the full workflow of prior
  construction, validation, and regulatory justification into a single
  package. We introduce bayprior, an R package and Shiny application that
  implements structured expert elicitation via quantile matching, moment
  matching, and the SHELF roulette method across six distribution
  families; linear and logarithmic expert opinion pooling with
  Bhattacharyya agreement diagnostics; prior-data conflict assessment
  using the Box p-value, surprise index, information divergence,
  Bhattacharyya overlap, and multivariate Mahalanobis distance;
  hyperparameter sensitivity analysis with tornado and influence heatmap
  visualisations; and robust, sceptical, and power prior alternatives. A
  regulatory reporting module generates self-contained HTML, PDF, or Word
  documents addressing expectations in the FDA's 2026 draft Bayesian
  methods guidance and in EMA guidance on incorporating external and
  historical information. The package is demonstrated on a synthetic
  oncology Phase II trial. bayprior is available on CRAN and includes a
  fully modular Shiny application for interactive use.
draft: true
author:
  - name: Ndoh Penn
    affiliation: Independent Researcher
    address:
    - Antwerp, Belgium
    url: https://github.com/ndohpenngit
    orcid: 0009-0003-9054-465X
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
header-includes:
  - \usepackage{float}
bibliography: RJreferences.bib
---

```{r setup, include=FALSE}
# print.bayprior() and related cli-based print methods fall back to plain
# cat() output when cli's styled rendering isn't available (e.g. no RSTUDIO/
# POSITRON env var and stdout isn't a tty, as in rmarkdown::render()'s callr
# subprocess) -- this fallback shipped in the v0.3.0 fix for #40. Setting
# RSTUDIO here instead forces the styled cli:: output path for this document,
# which is a cosmetic choice, not a requirement for output to appear at all.
Sys.setenv(RSTUDIO = "1")

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.pos    = "H",
  out.width  = "100%",
  comment    = "#>"
)
library(bayprior)
library(kableExtra)
```

# Introduction

Bayesian statistical methods are increasingly used in clinical trial design
and analysis, offering flexible incorporation of prior knowledge and natural
uncertainty quantification [@FDA2026]. However, the widespread adoption of
Bayesian methods in regulatory settings requires that prior distributions be
not only well-chosen but also transparently documented and demonstrably robust.
The FDA's 2026 draft guidance on Bayesian methods [@FDA2026] calls for
sponsors to demonstrate that trial conclusions hold under sensitivity
analyses over the prior specification, but does not prescribe specific
software tools or diagnostics for doing so.

In practice, prior specification often proceeds informally. Analysts may
elicit a prior from clinical literature or expert opinion without a
structured procedure, assess prior-data conflict informally by visual
inspection, and document the justification in narrative form. This approach
is difficult to reproduce and rarely satisfies regulatory reviewers who
expect a systematic, auditable justification.

Existing R packages for Bayesian clinical trials focus on posterior
computation and adaptive design. \CRANpkg{RBesT} [@RBesT] implements
robust meta-analytic-predictive priors for historical data borrowing.
\CRANpkg{trialr} [@trialr] provides a range of adaptive trial designs
via Stan. \CRANpkg{hdbayes} [@hdbayes] implements multiple historical
data borrowing priors -- including power priors, normalised power priors,
robust MAP priors, and commensurate priors -- for generalised linear
models. Each of these tools addresses a specific stage of the Bayesian
workflow well, but none carries a prior through construction from expert
knowledge, prior-data conflict assessment, and regulatory reporting as a
single connected process.

We introduce \CRANpkg{bayprior}, an R package and Shiny application that
connects these stages by providing an integrated prior justification
workflow: structured expert elicitation, expert opinion pooling,
prior-data conflict diagnostics, sensitivity analysis, robust and
regulatory prior alternatives, and automated regulatory reporting. The
package implements established statistical methodology [@OHagan2006;
@Box1980; @Oakley2010; @Schmidli2014; @Ibrahim2000; @Spiegelhalter1994]
in a unified, accessible framework suitable for both programmatic use and
interactive exploration via the bundled Shiny application.

# Related software

## Expert Elicitation

The Sheffield Elicitation Framework [@Oakley2010] provides a structured
approach to expert elicitation and is implemented in the R package
\CRANpkg{SHELF} [@SHELF]. \CRANpkg{SHELF} supports elicitation from
individual or multiple experts across a range of distribution families,
with weighted linear pooling for aggregating multiple expert beliefs.
\CRANpkg{bayprior} extends this with three programmatic elicitation
methods (quantile matching, moment matching, and the roulette method)
across six distribution families, adds logarithmic opinion pooling as
a complement to linear pooling, and introduces quantitative agreement
diagnostics via the Bhattacharyya coefficient for assessing
inter-expert consistency before pooling.

## Robust Priors and Historical Borrowing

\CRANpkg{RBesT} [@RBesT] implements the robust meta-analytic-predictive
(MAP) prior approach [@Schmidli2014] for incorporating historical control
data. \CRANpkg{hdbayes} [@hdbayes] provides a broader range of historical
borrowing methods, including power priors, normalised power priors,
commensurate priors, and robust MAP priors, for generalised linear models
via Stan. \CRANpkg{bayprior} provides the Schmidli robust mixture prior
and calibrated power priors [@Ibrahim2000] as alternatives within the prior
justification workflow, but focuses on prior specification and justification
rather than historical data meta-analysis.

## Prior Sensitivity and Conflict Diagnostics

\CRANpkg{priorsense} [@priorsense] implements power-scaling sensitivity
analysis for detecting prior-data conflict and likelihood noninformativity
in fitted Stan models [@Kallioinen2023]. \CRANpkg{bayprior} addresses a
complementary use case: closed-form conflict diagnostics for conjugate
prior families at the prior specification stage, before any model is
fitted, without requiring Stan. The two packages serve different points
in the Bayesian workflow.

## Scope Comparison

Table 1 summarises the scope of existing packages
relative to \CRANpkg{bayprior}.

```{r pkg-comparison, echo=FALSE}
knitr::kable(
  data.frame(
    Package     = c("SHELF", "RBesT", "trialr", "hdbayes",
                    "priorsense", "bayprior"),
    Elicitation = c("Yes",     "No",    "No",     "No",
                    "No",       "Yes"),
    Pooling     = c("Partial", "No",    "No",     "No",
                    "No",       "Yes"),
    Conflict    = c("No",      "No",    "No",     "No",
                    "Yes",      "Yes"),
    Sensitivity = c("No",      "No",    "No",     "No",
                    "Yes",      "Yes"),
    `Robust priors` = c("No",  "Yes",   "No",     "Yes",
                        "No",   "Yes"),
    Reporting   = c("No",      "No",    "No",     "No",
                    "No",       "Yes"),
    check.names = FALSE
  ),
  caption = "Scope comparison of R packages for Bayesian prior specification.
             Partial: SHELF pooling supports linear pooling only.
             priorsense conflict and sensitivity diagnostics require a
             fitted Stan model.",
  booktabs = TRUE
) %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

# Package overview

\CRANpkg{bayprior} is built using the \CRANpkg{golem} [@golem] framework
and provides both exported R functions for programmatic use and a modular
Shiny application for interactive exploration. The package is available on
CRAN and can be installed with:

```{r install, eval=FALSE}
install.packages("bayprior")

# Or the development version from GitHub:
pak::pak("ndohpenngit/bayprior")
```

The Shiny application can be launched with:

```{r run-app, eval=FALSE}
library(bayprior)
run_app()
```

## Prior Elicitation

\CRANpkg{bayprior} implements three structured elicitation approaches
across six distribution families (Beta, Normal, Gamma, Log-Normal,
Exponential, and Weibull).

**Moment matching** derives hyperparameters analytically from an
expert-supplied mean and standard deviation:

```{r elicitation-moments}
prior <- elicit_beta(
  mean      = 0.35,
  sd        = 0.10,
  method    = "moments",
  label     = "Response rate",
  expert_id = "Expert_1"
)
```

```{r prior-table, echo=FALSE}
knitr::kable(
  data.frame(
    Quantity = c("Distribution", "alpha", "beta", "Method", "Expert",
                  "Mean (SD)", "95% CrI"),
    Value = c(
      toupper(prior$dist),
      round(prior$params$alpha, 4),
      round(prior$params$beta, 4),
      paste(prior$method, "elicitation"),
      prior$expert_id,
      paste0(round(prior$fit_summary$mean, 4), " (",
             round(prior$fit_summary$sd, 4), ")"),
      paste0("[", round(prior$fit_summary$q025, 4), ", ",
             round(prior$fit_summary$q975, 4), "]")
    )
  ),
  caption = "Elicited Beta prior for the response rate (moment matching)."
) %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

**Quantile matching** fits the distribution numerically to expert-specified
probability--value pairs, accommodating asymmetric beliefs that moment
matching cannot capture. For example, an expert who believes the response
rate has a 25\% chance of being below 0.20, a 60\% chance of being below
0.35, and an 80\% chance of being below 0.50:

```{r elicitation-quantile}
prior_qm <- elicit_beta(
  quantiles = c(`0.25` = 0.20, `0.60` = 0.35, `0.80` = 0.50),
  method    = "quantile",
  label     = "Response rate (quantile)",
  expert_id = "Expert_1"
)
```

```{r elicitation-plot, fig.cap="Beta prior elicited by moment matching (mean = 0.35, SD = 0.10). Shaded region shows the 95\\% credible interval."}
plot(prior)
```

The fitted distribution, parameter estimates, and a 95\% credible interval
are displayed automatically. All six families share a common API, so
`elicit_beta()` can be replaced with `elicit_normal()`, `elicit_gamma()`,
`elicit_lognormal()`, `elicit_exponential()`, or `elicit_weibull()` with
identical syntax.

## Expert Pooling

Multiple expert priors can be aggregated using linear or logarithmic
opinion pooling [@OHagan2006]. Linear pooling takes the weighted mixture
of the densities, $\pi(\theta) = \sum_k w_k \pi_k(\theta)$; logarithmic
pooling takes the weighted geometric mean, $\pi(\theta) \propto
\prod_k \pi_k(\theta)^{w_k}$. For linear pooling, the consensus mean and
SD are computed exactly from the component means/SDs and weights, without
numerical integration:
$$\text{mean} = \sum_k w_k \bar{x}_k, \qquad
  \text{var} = \sum_k w_k \left(s_k^2 + \bar{x}_k^2\right) - \text{mean}^2$$
No closed-form SD exists for logarithmic pooling in general.

```{r pooling}
e1 <- elicit_beta(mean = 0.30, sd = 0.08, method = "moments",
                  label = "Response rate", expert_id = "Expert_1")
e2 <- elicit_beta(mean = 0.42, sd = 0.10, method = "moments",
                  label = "Response rate", expert_id = "Expert_2")

consensus <- aggregate_experts(
  priors  = list(E1 = e1, E2 = e2),
  weights = c(0.5, 0.5),
  method  = "linear"
)
```

```{r consensus-table, echo=FALSE}
knitr::kable(
  data.frame(
    Quantity = c("Distribution", "Components", "Weights", "Mean (SD)"),
    Value = c(
      toupper(consensus$dist),
      length(consensus$components),
      paste(round(consensus$weights, 3), collapse = ", "),
      paste0(round(consensus$fit_summary$mean, 4), " (",
             round(consensus$fit_summary$sd, 4), ")")
    )
  ),
  caption = "Consensus prior from linear pooling of two experts."
) %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

Agreement between experts can be quantified by the Bhattacharyya
coefficient [@Bhattacharyya1943], which measures distributional overlap on
a scale from 0 (no overlap) to 1 (identical distributions). A coefficient
close to 1 supports pooling; a low coefficient indicates the experts'
beliefs are largely non-overlapping, in which case pooling may obscure
a genuine disagreement that should be reported separately.

## Prior-Data Conflict Diagnostics

Prior-data conflict occurs when the observed data is implausible under the
prior predictive distribution [@Box1980]. \CRANpkg{bayprior} computes four
complementary conflict metrics using a common Normal approximation to the
prior and to the observed-data likelihood (via their means and standard
errors), so that the same diagnostics apply uniformly whatever the prior's
distribution family or the data type:

- **Prior predictive p-value** [@Box1980]: the probability of observing
  data at least as extreme as the observed data under the prior predictive
  distribution.
- **Surprise index**: the standardised distance between the prior mean and
  the observed data, expressed in prior standard deviation units.
- **Information divergence**: the Kullback-Leibler divergence from the
  prior to the normalised likelihood.
- **Bhattacharyya overlap**: the distributional overlap between the prior
  and normalised likelihood.

```{r conflict}
interim <- list(type = "binary", x = 18, n = 40)
cd <- prior_conflict(prior, interim)
```

```{r cd-table, echo=FALSE}
knitr::kable(
  data.frame(
    Diagnostic = c("Box's p-value", "Surprise index", "KL divergence",
                    "Bhattacharyya overlap", "Conflict severity",
                    "Recommendation"),
    Value = c(
      round(cd$box_pvalue, 4),
      round(cd$surprise_index, 4),
      round(cd$kl_prior_likelihood, 4),
      round(cd$overlap, 4),
      toupper(cd$conflict_severity),
      cd$recommendation
    )
  ),
  caption = "Prior-data conflict diagnostics for the response rate prior.",
  booktabs = TRUE, linesep = ""
) %>%
  kableExtra::column_spec(1, width = "3.5cm") %>%
  kableExtra::column_spec(2, width = "9cm") %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

Four data types are supported for the observed-data side of this
comparison: binary (event count out of a sample size), continuous
(observed mean and SD), Poisson/count (events over exposure), and survival
(events over total follow-up time); each is mapped internally to an
approximate mean and standard error before the Normal-approximation
diagnostics above are applied. This is distinct from the *exact* conjugate
posterior updates (Beta-Binomial, Normal-Normal, Gamma-Poisson,
Gamma-Exponential) that \CRANpkg{bayprior} uses elsewhere -- for example
in the sensitivity analysis module below -- to compute posterior summaries.

For co-primary endpoints, multivariate conflict can be assessed using the
Mahalanobis distance [@Mahalanobis1936]. The test assumes approximate
multivariate normality; for proportion endpoints such as the illustrative
example below, `bayprior`'s documentation recommends transforming to the
log-odds scale first. For simplicity, the example here uses raw
proportions directly:

```{r conflict-mv}
pm   <- c(0.35, 0.60)
pcov <- matrix(c(0.010, 0.003, 0.003, 0.015), 2, 2)
om   <- c(0.52, 0.58)
ocov <- matrix(c(2e-4, 4e-5, 4e-5, 2e-4), 2, 2)

mv <- conflict_mahalanobis(pm, pcov, om, ocov,
                           labels = c("Response rate", "OS rate"))
```

```{r mv-table, echo=FALSE}
knitr::kable(
  data.frame(
    Diagnostic = c("Mahalanobis D", "Chi-sq p-value (df=2)", "Conflict flag",
                    paste("z -", names(mv$marginal_z)), "Interpretation"),
    Value = c(
      round(mv$mahal_distance, 4),
      round(mv$pvalue, 4),
      as.character(mv$conflict_flag),
      round(mv$marginal_z, 3),
      mv$interpretation
    )
  ),
  caption = "Multivariate Mahalanobis conflict check for two co-primary endpoints.",
  booktabs = TRUE, linesep = ""
) %>%
  kableExtra::column_spec(1, width = "3.5cm") %>%
  kableExtra::column_spec(2, width = "9cm") %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

## Sensitivity Analysis

The sensitivity module evaluates posterior conclusions across a
hyperparameter grid, answering the regulatory question: does the
treatment benefit conclusion change materially under plausible alternative
priors?

```{r sensitivity, fig.cap="Tornado plot showing the influence of prior hyperparameters on the posterior mean. Parameters are ranked by their marginal influence."}
sa <- sensitivity_grid(
  prior        = prior,
  data_summary = interim,
  param_grid   = list(
    alpha = seq(1, 8, by = 0.5),
    beta  = seq(2, 20, by = 1)
  ),
  target    = c("posterior_mean", "prob_efficacy"),
  threshold = 0.30
)
plot_tornado(sa)
```

The `sensitivity_cri()` function tracks credible interval width
specifically -- a quantity directly relevant to the precision of treatment
effect estimates in regulatory submissions.

## Robust, Sceptical, and Power Priors

\CRANpkg{bayprior} provides three alternative prior types for regulatory
sensitivity analyses.

**Robust mixture prior** [@Schmidli2014]: mixes the informative prior with
a vague component, protecting against prior misspecification:

```{r robust}
rob <- robust_prior(prior, vague_weight = 0.20)
plot(rob)
```

**Sceptical prior** [@Spiegelhalter1994]: centred at the null value of
the treatment effect, representing a conservative regulatory stance:

```{r sceptical}
scep <- sceptical_prior(
  null_value = 0.20,
  family     = "beta",
  strength   = "moderate",
  label      = "Response rate (sceptical)"
)
```

**Power prior** [@Ibrahim2000]: down-weights historical data by a factor
$\delta \in (0, 1]$ calibrated to achieve a target Bayes Factor:

```{r power-prior}
base  <- elicit_beta(mean = 0.50, sd = 0.20, method = "moments",
                     label = "Response rate")
calib <- calibrate_power_prior(
  historical_data = list(type = "binary", x = 12, n = 40),
  current_data    = list(type = "binary", x = 18, n = 50),
  base_prior      = base,
  target_bf       = 3
)
```

```{r calib-table, echo=FALSE}
knitr::kable(
  data.frame(
    Quantity = c("Method", "Target BF", "Optimal delta",
                  "Power prior mean", "Power prior SD"),
    Value = c(
      calib$method,
      calib$target_bf,
      calib$delta_opt,
      round(calib$power_prior$fit_summary$mean, 4),
      round(calib$power_prior$fit_summary$sd, 4)
    )
  ),
  caption = "Power prior calibration via Bayes Factor."
) %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

## Regulatory Reporting

A single call to `prior_report()` generates a self-contained prior
justification document in HTML, PDF, or Word format:

```{r report, eval=FALSE}
prior_report(
  prior           = prior,
  conflict        = cd,
  sensitivity     = sa,
  robust_prior    = rob,
  sceptical_prior = scep,
  output_format   = "html",
  trial_name      = "TRIAL-001",
  sponsor         = "Example Pharma Ltd",
  author          = "N.P., Biostatistician"
)
```

The generated report includes parameter summary tables, all diagnostic
plots, a compliance checklist aligned with the FDA 2026 draft guidance
[@FDA2026], and a structured prior justification narrative.

# Case study: TRIAL-001

We demonstrate the \CRANpkg{bayprior} workflow end to end on TRIAL-001, a
synthetic oncology Phase II trial with a binary response rate endpoint.
The target response rate is 30\% (null hypothesis) versus an expected
rate of 40\% (alternative hypothesis). TRIAL-001 is single-arm with no
historical control data, so the power prior module is not demonstrated
here; a worked example appears in the Package overview section above.

## Expert Elicitation and Pooling

Two clinical experts independently specified Beta priors for the response
rate. Expert 1 (conservative) placed a mean of 0.30 with SD of 0.08;
Expert 2 (optimistic) placed a mean of 0.42 with SD of 0.10. Equal weights
were assigned on the basis of comparable seniority.

```{r trial-elicitation}
e1 <- elicit_beta(mean = 0.30, sd = 0.08, method = "moments",
                  label = "Response rate", expert_id = "Expert_1")
e2 <- elicit_beta(mean = 0.42, sd = 0.10, method = "moments",
                  label = "Response rate", expert_id = "Expert_2")

pool <- aggregate_experts(list(E1 = e1, E2 = e2),
                          weights = c(0.5, 0.5),
                          method  = "linear")
```

```{r pool-table, echo=FALSE}
knitr::kable(
  data.frame(
    Quantity = c("Distribution", "Components", "Weights", "Mean (SD)"),
    Value = c(
      toupper(pool$dist),
      length(pool$components),
      paste(round(pool$weights, 3), collapse = ", "),
      paste0(round(pool$fit_summary$mean, 4), " (",
             round(pool$fit_summary$sd, 4), ")")
    )
  ),
  caption = "Consensus prior pooled from two clinical experts (TRIAL-001)."
) %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

The two expert distributions overlap substantially (means 0.30 and
0.42 with comparable SDs of 0.08 and 0.10), supporting the use of
equal-weight linear pooling. The consensus prior had a mean of
`r round(pool$fit_summary$mean, 3)` and SD of
`r round(pool$fit_summary$sd, 3)`.

## Conflict Diagnostics at Interim

At the pre-specified interim analysis, 18 responses were observed in
40 patients (observed rate: 45\%). Following common practice for Box's
prior predictive check, a significance threshold of $\alpha = 0.05$ was
pre-specified: a Box p-value below this threshold would indicate evidence
of prior-data conflict warranting reassessment of the prior. This is the
default threshold in `prior_conflict()`.

```{r trial-conflict}
interim_data <- list(type = "binary", x = 18, n = 40)
cd_trial <- prior_conflict(pool, interim_data, alpha = 0.05)
```

```{r cdtrial-table, echo=FALSE}
knitr::kable(
  data.frame(
    Diagnostic = c("Box's p-value", "Surprise index", "KL divergence",
                    "Bhattacharyya overlap", "Conflict severity",
                    "Recommendation"),
    Value = c(
      round(cd_trial$box_pvalue, 4),
      round(cd_trial$surprise_index, 4),
      round(cd_trial$kl_prior_likelihood, 4),
      round(cd_trial$overlap, 4),
      toupper(cd_trial$conflict_severity),
      cd_trial$recommendation
    )
  ),
  caption = "Prior-data conflict diagnostics at the TRIAL-001 interim analysis.",
  booktabs = TRUE, linesep = ""
) %>%
  kableExtra::column_spec(1, width = "3.5cm") %>%
  kableExtra::column_spec(2, width = "9cm") %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

The Box p-value was above the pre-specified 0.05 threshold, indicating no
evidence of prior-data conflict. The posterior distribution updated the
consensus prior towards the observed data as expected.

## Enthusiastic and Sceptical Priors

A common regulatory sensitivity practice, following Spiegelhalter and
Freedman (1994), is to present conclusions under both an enthusiastic
prior (favouring treatment benefit) and a sceptical prior (centred at the
null), alongside the consensus prior already assessed above:

```{r trial-sensitivity-pair}
enth_trial <- elicit_beta(mean = 0.45, sd = 0.08, method = "moments",
                          label = "Response rate (enthusiastic)")
scep_trial <- sceptical_prior(null_value = 0.20, family = "beta",
                              strength = "moderate",
                              label = "Response rate (sceptical)")

cd_enth <- prior_conflict(enth_trial, interim_data, alpha = 0.05)
cd_scep <- prior_conflict(scep_trial, interim_data, alpha = 0.05)

cat("Conflict severity (consensus):    ", cd_trial$conflict_severity, "\n")
cat("Conflict severity (enthusiastic): ", cd_enth$conflict_severity, "\n")
cat("Conflict severity (sceptical):    ", cd_scep$conflict_severity, "\n")
```

Conclusions were consistent across all three priors, supporting the
robustness of the trial's conclusions to reasonable variation in the
prior specification.

## Sensitivity Analysis

Beyond comparing discrete alternative priors, the sensitivity module
evaluates how the posterior conclusion changes continuously across a
hyperparameter grid. Because the grid is defined over the hyperparameters
of a single distribution family, `sensitivity_grid()` moment-matches a
mixture (pooled) prior's actual blended mean and SD to a working Beta
prior before gridding, rather than analysing a single expert's prior in
isolation; a message reports the working prior used (here, mean
$\approx$ 0.36, SD $\approx$ 0.11, reflecting both experts):

```{r trial-sensitivity, fig.cap="Tornado plot for TRIAL-001: influence of the moment-matched pooled prior's Beta hyperparameters on the posterior mean, given the interim data."}
sa_trial <- sensitivity_grid(
  prior        = pool,
  data_summary = interim_data,
  param_grid   = list(
    alpha = seq(3, 12, by = 1),
    beta  = seq(6, 20, by = 1)
  ),
  target    = c("posterior_mean", "prob_efficacy"),
  threshold = 0.30
)
plot_tornado(sa_trial)
```

The posterior mean and the probability of exceeding the 30\% efficacy
threshold both remained stable across the explored hyperparameter range,
indicating the trial's conclusion is not driven by fine details of the
pooled prior's specification.

## Robust Mixture Prior

As a further regulatory sensitivity check, a robust mixture prior mixes
the consensus prior with a vague component, bounding the influence any
single prior specification can have on the posterior:

```{r trial-robust}
rob_trial <- robust_prior(pool, vague_weight = 0.20)
cd_rob    <- prior_conflict(rob_trial, interim_data, alpha = 0.05)

cat("Conflict severity (robust mixture):", cd_rob$conflict_severity, "\n")
```

The robust mixture prior showed the same conflict conclusion as the
consensus, enthusiastic, and sceptical priors above.

## Regulatory Report

The full set of TRIAL-001 analyses -- the pooled prior, conflict
diagnostics, sensitivity analysis, and robust and sceptical alternatives
-- can be assembled into a single submission-ready document with one
function call:

```{r trial-report, eval=FALSE}
prior_report(
  prior           = pool,
  conflict        = cd_trial,
  sensitivity     = sa_trial,
  robust_prior    = rob_trial,
  sceptical_prior = scep_trial,
  output_format   = "html",
  trial_name      = "TRIAL-001",
  sponsor         = "Example Pharma Ltd",
  author          = "N.P., Biostatistician"
)
```

# Discussion

\CRANpkg{bayprior} provides a structured workflow for prior justification
that addresses a practical gap in the Bayesian clinical trial toolkit. The
package is designed to be used alongside existing analysis packages
(\CRANpkg{RBesT}, \CRANpkg{rstanarm}, \CRANpkg{brms}) rather than
replacing them: bayprior handles the prior specification phase, and
established packages handle the posterior computation.

## Regulatory Alignment

The FDA's 2026 draft guidance on Bayesian methods [@FDA2026] asks sponsors
relying on a Bayesian analysis to provide detailed support for the
proposed prior distribution and any external information borrowing, to
quantify the prior's influence (for example via an effective sample
size), to evaluate operating characteristics -- including sensitivity to
the prior specification and to prior-data conflict, typically via
simulation -- and to document the analysis in sufficient detail for the
FDA to understand and, where feasible, reproduce it. \CRANpkg{bayprior}
supports each of these expectations directly: structured elicitation and
pooling provide a documented, reproducible derivation of the prior; the
sensitivity, robust, sceptical, and power prior modules provide alternative
specifications for demonstrating robustness; and the regulatory reporting
module assembles this into a single submission-ready document.

Prior-data conflict and the influence of external information are also a
recurring theme in EMA guidance on incorporating historical or external
data. The EMA reflection paper on extrapolation in paediatric drug
development [@EMA2018], for instance, requires sponsors to quantify how
much information a Bayesian prior contributes relative to the data
generated in the target population and to investigate the Type I error
properties of the resulting analysis. The conflict diagnostics in
\CRANpkg{bayprior} -- the Box predictive p-value, surprise index,
information divergence, and Bhattacharyya overlap -- provide a closed-form
way to address this for conjugate prior families, without requiring a
fitted model.

## Shiny Application

The bundled Shiny application mirrors the package's six modules as
separate panels, allowing a biostatistician without R programming
experience to perform the same elicitation, pooling, conflict, sensitivity,
robust prior, and reporting workflow interactively. Each panel produces the
same plots and summary statistics as the corresponding R function, and the
final regulatory report can be downloaded directly from the application.
The application is hosted at
<https://ndohpenn-bayprior.share.connect.posit.cloud> and can be launched
locally via `run_app()`.

## Limitations

Several limitations apply to the current release (v0.3.1). The
multivariate Mahalanobis check is currently limited to bivariate (k = 2)
endpoints; extension to k $\geq$ 3 co-primary endpoints is planned for a
future release. Survival power priors require numerical integration over
the partial likelihood and are not yet implemented; calibrated power priors
are currently available for binary, continuous, Poisson, and log-normal
data types. The robust mixture prior uses a Normal vague component
regardless of the informative prior family; for Beta, Gamma, and other
non-Normal informative priors, the resulting mixture density is computed
numerically rather than analytically, which is accurate in the body of the
distribution but may be less reliable in the extreme tails. Sensitivity
grids over mixture priors moment-match the pooled mean and SD to a
working prior in the dominant component's family (Beta, Normal, Gamma, or
Log-Normal); families without a mean/SD parameterisation (Exponential,
Weibull) fall back to the dominant component alone, with a warning.

Future development includes survival power priors, k $\geq$ 3 endpoint
Mahalanobis conflict, and interoperability with \CRANpkg{RBesT} for MAP
prior workflows. A companion paper in Pharmaceutical Statistics
(in preparation) provides a methodological treatment of the conflict
diagnostic framework and a regulatory alignment assessment against the
FDA 2026 draft guidance.

# Summary

\CRANpkg{bayprior} provides an integrated R tool for the complete prior
justification workflow in Bayesian clinical trials, combining structured
expert elicitation, opinion pooling, conjugate conflict diagnostics,
hyperparameter sensitivity analysis, robust prior alternatives, and
regulatory reporting in a single package. The package is available on CRAN:

```{r install-cran, eval=FALSE}
install.packages("bayprior")
```

The full documentation and vignettes are available at
<https://ndohpenngit.github.io/bayprior/>. A live Shiny application
is hosted at <https://ndohpenn-bayprior.share.connect.posit.cloud>.

```{r citation, echo=FALSE}
cat(format(citation("bayprior"), style = "text"), sep = "\n")
```
